Complex aneuploidy triggers autophagy and p53-mediated apoptosis and impairs the second lineage segregation in human preimplantation embryos

Version 1

Abstract

Abstract

About 70% of human cleavage stage embryos show chromosomal mosaicism, falling to 20% in blastocysts. Chromosomally mosaic human blastocysts can implant and lead to healthy new-borns with normal karyotypes. Studies in mouse embryos and human gastruloids have shown that aneuploid cells show proteotoxic stress, autophagy and p53 activation and that they are eliminated from the epiblast by apoptosis while being rather tolerated in the trophectoderm. These observations suggest a selective loss of aneuploid cells from human embryos, but the underlying mechanisms are not yet fully understood. In this study we investigated the cellular consequences of aneuploidy in a total of 85 human blastocysts. RNA-sequencing of trophectoderm cells showed transcriptional signatures of a deregulated p53 pathway and apoptosis, which was proportionate to the level of chromosomal imbalance. Immunostaining revealed that aneuploidy triggers proteotoxic stress, autophagy and apoptosis in aneuploid embryos. Total cell numbers were lower in aneuploid embryos, due to a decline both in trophectoderm and in epiblast/primitive endoderm cell numbers. While lower cell numbers in trophectoderm may be attributed to apoptosis, it appeared that aneuploidy impairs the second lineage segregation and primitive endoderm formation in particular. Our findings might explain why fully aneuploid embryos fail to further develop and we hypothesize that the same mechanisms lead to removal of aneuploid cells from mosaic embryos. This hypothesis needs further study as we did not analyse chromosomal mosaic embryos. Finally, we demonstrated clear differences with previous findings in the mouse, emphasizing the need for human embryo research to understand the consequences of aneuploidy.

Introduction

Aneuploidy is common in human preimplantation embryos from both natural (Munné et al., 2020a) and medically assisted reproduction cycles (Voullaire et al., 2000; Wells and Delhanty, 2000; Vanneste et al., 2009; Johnson et al., 2010; Chavez et al., 2012; Mertzanidou et al., 2013b, 2013a; Chow et al., 2014; Fragouli et al., 2019; Popovic et al., 2020; Starostik et al., 2020; Capalbo et al., 2021; Yang et al., 2021). Depending on the study, up to 80% of human cleavage stage embryos are found to be aneuploid, of which 70% show chromosomal mosaicism, i.e. the presence of at least two cell lineages with different genomic content (Regin et al., 2022). The reasons behind this high rate of chromosomal abnormalities of mitotic origin is still largely unknown (McCoy, 2017) but could be a combination of a weak spindle assembly checkpoint (SAC) that is uncoupled of apoptosis until the blastocyst stage (Jacobs et al., 2017; Vázquez-Diez et al., 2019), in combination with an abundancy of transcripts of anti-apoptotic genes in cleavage-stage embryos (Yan et al., 2013). After embryonic genome activation, the rapidly increasing expression of pro-apoptotic genes along with the full functionality of the SAC establishes the control of mitotic errors.

It is by now well established that chromosomally mosaic embryos can result in healthy new-borns with normal karyotypes, implying a progressive and selective loss of aneuploid cells during development (Greco et al., 2015; Fragouli et al., 2017, 2019; Munné et al., 2017, 2020b; Capalbo et al., 2021; Viotti et al., 2021; Yang et al., 2021). In the preimplantation embryo, several studies have shown that the proportion of aneuploid cells within the embryo decreases three days post fertilization (3dpf) (van Echten-Arends et al., 2011; Fragouli et al., 2019; Yang et al., 2021) with no apparent preferential allocation of aneuploid cells during the first lineage segregation event to either trophectoderm (TE) or inner cell mass (ICM) (Capalbo et al., 2013; Popovic et al., 2018; Starostik et al., 2020; Ren et al., 2022). However, aneuploid cells can be excluded as cell debris (Orvieto et al., 2020) or allocated to the blastocoel cavity and to peripheral cells that do not participate in the formation of the embryo (Griffin et al., 2022). In some studies, (Starostik et al., 2020; Griffin et al., 2022) the TE contained a slightly higher number of aneuploid cells than the ICM, without reaching statistical significance. In their landmark paper, (Bolton et al., 2016) found a quite similar but statistically significant 6% TE enrichment in mouse embryos treated with the SAC inhibitor reversine to induce mosaic aneuploidy. This small enrichment fell in the same confidence interval as the Starostik study, therefore TE enrichment cannot be ruled out in the human embryo. After implantation, aneuploid cells become progressively more frequent in the trophoblast lineage which is derived from TE cells as compared to the epiblast (EPI) and primitive endoderm (PrE) lineages which are derived from the ICM cells during the second lineage segregation (Starostik et al., 2020). This suggests that the trophoblast cells are more tolerant to aneuploidy, in line with the rate of 1-2% of confined placental mosaicism observed in ongoing pregnancies in the general population (Kalousek and Vekemans, 1996) and with the presence of aneuploid cells in the placental lineages only (Zamani Esteki et al., 2019). This is further supported by a recent study showing that genetic mosaicism is a normal feature of the placenta and is due to genetic bottlenecks that occur early in development (Coorens et al., 2021). Taken together, strong evidence supports the selective elimination of aneuploid cells from the embryonic lineage in the pre- and post-implantation development, while the placenta appears to be more tolerant to genetic imbalances.

From a cellular point of view, aneuploidy has two types of consequences. Imbalances of a specific chromosome, such as trisomy 21 or trisomy 18, can result in different phenotypes, as evidenced by the human Down and Edwards syndromes. These imbalances can manifest early on during early embryo development (Licciardi et al., 2018; Fuchs Weizman et al., 2019; Sanchez-Ribas et al., 2019; Shahbazi et al., 2020). Moreover, aneuploidy as such also elicits two highly conserved intracellular stress responses, that have been found from yeast and plants to mouse and human cells (reviewed in: Santaguida and Amon, 2015; Zhu et al., 2018; Chunduri and Storchová, 2019; Chunduri et al., 2022; Krivega et al., 2022), and which are likely at the basis of the selective clearance of aneuploid cells during development. First, aneuploidy results in gene-dosage defects, leading to an unbalanced protein pool and proteotoxic stress (Torres et al., 2007; Huettel et al., 2008; Williams et al., 2008; Tang et al., 2011; Oromendia et al., 2012; Stingele et al., 2012; Donnelly et al., 2014; Ohashi et al., 2015). If this remains unresolved by activation of autophagy (Stingele et al., 2012; Dürrbaum et al., 2014; Santaguida et al., 2015; Ariyoshi et al., 2016), CASP8/LC3B/p62 may interact to activate the apoptosis cascade (Pan et al., 2011). Second, aneuploidy results in replication stress that can lead to DNA-damage, resulting in p53 activation (Li et al., 2010; Thompson and Compton, 2010; Santaguida et al., 2017) and ultimately to cell cycle arrest and apoptosis (reviewed in: Santaguida and Amon, 2015; Zhu et al., 2018; Regin et al., 2022).

During early embryo development of the mouse, Bolton et al. (Bolton et al., 2016) showed that aneuploid cells are selectively cleared from chimeric embryos generated from control blastomeres mixed with aneuploid blastomeres obtained using a treatment with the SAC inhibitor reversine. Later work by the same group showed that apoptosis of these aneuploid cells requires autophagy and p53 activation and that euploid cells in the mouse embryo compensate this cell loss by increased proliferation (Singla et al., 2020). The authors proposed that the proteotoxic stress in these aneuploid chimeric mouse embryos activated the p53 pathway, which in turn activated autophagy and apoptosis. In a recent study using human 2D gastruloids with reversine-induced aneuploidy, apoptosis was only induced in differentiating but not in pluripotent cells, although aneuploidy induced an increase in total nuclear p53 proteins (Yang et al., 2021). The apoptotic response to aneuploidy also seems to be lineage-specific. In the mouse, aneuploid cells in the ICM are eliminated by apoptosis while they are rather tolerated in the TE (Bolton et al., 2016). Furthermore, in a follow-up study, (Singla et al., 2020) demonstrated that during the peri-implantation period, aneuploid cells were eliminated from the EPI and the PrE at similar rates, and that the EPI-specific elimination of aneuploid cells by apoptosis is coupled to increased proliferation rates of euploid cells. Similarly, in human gastruloids, aneuploid cells are eliminated by apoptosis from the embryonic germ layers but are tolerated in the TE-like cells (Yang et al., 2021).

The type of aneuploidy and the percentage of aneuploid cells within a euploid/aneuploid mosaic embryo determine its developmental capacity. The viability of mouse embryos inversely correlates to the proportion of aneuploid cells, with 50% of euploid cells being sufficient to sustain normal development (Bolton et al., 2016). This same threshold appears to apply to human embryos, where embryos with less than 50% of aneuploid cells have higher developmental potential than those containing more than 50%, albeit both performing worse than fully euploid embryos (Viotti et al., 2021, Capalbo et al., 2021). Conversely, whole chromosome gains and losses that are present in every cell of the human embryo are usually linked to lethality with the notable exception of trisomies 13 (Patau syndrome), 18 (Edwards syndrome) and 21 (Down syndrome), and sex chromosome abnormalities. Shahbazi et al. recently described the consequences of carrying specific meiotic aneuploidies on human embryo development using an in vitro implantation model (Shahbazi et al., 2020). They showed that human embryos with trisomy 15, 16 and 21 and monosomy 21 readily all reached the blastocyst stage, and that monosomy 21 embryos arrested after implantation while trisomy 15 and 21 embryos developed further. Interestingly, trisomy 16 embryos showed a hypoproliferative trophoblast due to excessive CDH1 expression, a gene located on chromosome 16.

In this study, we investigated the cellular responses to complex aneuploidy during human preimplantation development, with a focus on stress response and lineage segregation events.

Methods

Ethical approval

All experiments have been approved by the local Commission of Medical Ethics of the UZ Brussel (B.U.N. 143201628722) and the Federal Committee for Medical and Scientific Research on Human Embryos in vitro (AdV069 and AdV091). Patients from Brussels IVF (UZ Brussel) donated their embryos for research after written informed consent.

Culture of human pre-implantation embryos

We warmed vitrified human blastocysts (5-6dpf) after PGT or at 3dpf using the Vitrification Thaw Kit (Vit Kit-Thaw; Irvine Scientific, USA) following manufacturer’s instructions. Embryos were graded before vitrification by experienced clinical embryologists according to Gardner and Schoolcraft criteria (Gardner and Schoolcraft, 1999). Post PGT-embryos were left to recover for 2h in 25 µL droplets of Origio Sequential BlastTM medium (Origio) at 37°C with 5% O2, 6% CO2 and 89% N2, after which they underwent a second TE biopsy (Fig. 1a). For the reversine experiment, we cultured the 3dpf embryos for 24h in the same medium supplemented with 0.5 µM reversine (Stem Cell Technologies) and after wash-out we let them develop until 5dpf and subsequently took a biopsy (Fig. 1b). The PGT embryos (5dpf or 6dpf) used for the second experiment series (Fig. 4a) were cultured for 16h after warming to ensure sufficient time to progress to the second lineage differentiation. Prior to fixation we live-stained all the embryos with either Caspase-3/7 or Caspase-8 (Supplementary Table 30) for 30 min and subsequently removed the zona pellucida using Acidic Tyrode’s Solution (Sigma-Aldrich) for 30 sec – 2 min and subsequently washed in PBS/2%BSA.

RNA-sequencing of trophectoderm cells reveals deregulation of DNA damage, p53, cell-cycle, autophagy and apoptosis pathways in chromosomally abnormal human embryos.

a. Human blastocysts 5 or 6 days post-fertilization (dpf) underwent Preimplantation Genetic Testing (PGT) for monogenic diseases and/or aneuploidy screening, followed by cryopreservation. Embryos that were not transferred were donated for research after written informed content. After warming and 2hrs recuperation, euploid and fully aneuploid research embryos underwent a second TE biopsy for RNA-sequencing. The remainder of the embryos was live-stained with either CASP3/7 or CASP8 for 30min and fixed for immunostaining for proteins of interest (see figure 3). b. 3dpf human embryos were warmed and treated with 0.5µM reversine for 24hrs then transferred to normal culture medium until 5dpf to undergo TE biopsy for RNA-sequencing. After 2hrs recuperation the embryos were live-stained with CASP8 for 30min and fixed for immunostaining for proteins of interest (see figure 3). c. Unsupervised hierarchical clustering of all coding genes. Color coding: beige = euploid, red = aneuploid and blue = reversine-treated. d. Comparison between the diagnosis of aneuploid embryos obtained after PGT (list on the left) and the results obtained after using InferCNV after RNA-sequencing (plot on the right). e,f. Volcano plots after differential gene expression analysis with a cutoff value of |log2 fold change| > 1 and −log10(p-value) < 0.05 for aneuploid versus euploid (e) and reversine versus euploid (f). g,h. Venn diagrams comparing the upregulated (g) and downregulated (h) genes in aneuploid versus reversine treated embryos. i,j. Supervised plot of pathway analysis using the Molecular Signatures Database (FDR < 0.05) based on the list of all differentially expressed genes between aneuploid or reversine versus euploid embryos. Pathways are ranked based on the total number of genes in overlap of both test sets, ranked from sets with the highest number of genes in overlap to sets with the lowest. Outputs of the Hallmark library and C5 ontology gene sets are shown in (i) and (j), respectively. k,l. Supervised plots after Ingenuity Pathway Analysis of upstream regulators with an activation z-score ≤ −2 and ≥ 2 and a p-value < 0.05 in the aneuploid (k) or reversine (l) versus euploid embryos. m,n. Supervised plots after Ingenuity Pathway Analysis of canonical pathways in aneuploid (m) or reversine (n) versus euploid embryos. Pathways are ranked by p-value from the most to the least significant, those with a p-value < 0.05 and with an activation z-score ≤ −1.5 and ≥ 1.5 were considered as inhibited or activated, respectively. Significant pathways with an activation z-score between −1.5 and 1.5 were considered as deregulated.

Biopsy procedure

The zona pellucida was opened (15-25µm) on 4dpf using a Saturn Laser (Cooper Surgical). For PGT cycles, expanded good quality embryos were biopsied on 5dpf; early blastocysts were evaluated again for expansion on 6dpf. Briefly, embryos were transferred to HEPES buffered medium and biopsied on a Nikon inversed microscope equipped with a Saturn Laser. The herniated TE cells were then aspirated and cut using manual laser shoots. In case of sticking cells this was supported by mechanical cutting. After successful tubing, the blastocysts were vitrified using CBS-VIT High-Security straws (CryoBioSystem, L’Aigle, France) with dimethylsulphoxide-ethylene glycol-sucrose as the cryoprotectants (Irvine Scientific Freeze Kit, Newtownmountkennedy, County Wicklow, Ireland) (Van Landuyt et al., 2011). The TE biopsy for RNA-sequencing (Fig. 1a,b) was performed with the same procedure, cells were biopsied form the same opening in the zona pellucida.

PGT

During PGT we analyzed a TE sample with multiple cells for each embryo. Next Generation Sequencing: The chromosomal analysis was performed by WGA (Sureplex, Illumina) followed by library preparation (KAPA HyperPlus, Roche), sequencing on a NovaSeq (Illumina) and an in-house developed interpretation pipeline. The analysis has an effective resolution of 5 Mb.

SNP-array: We relied on whole genome amplification (MDA, Repli-G) followed by genome wide SNP array using Karyomapping (Vitrolife) with Bluefuse software (Vitrolife). In addition, SNP array data were analyzed with an in-house developed interpretation pipeline for aneuploidy detection.

RNA-sequencing

Multiple (5-10) TE cells were used as input to generate full length cDNA with the SMART-SeqTM v4 UltraTM Low Input RNA Kit (Clontech Laboratories, Inc.) according to the manufacturer’s instructions. The quality of cDNA was checked using the AATI Fragment Analyzer (Advances Analytical). Library preparation was performed using the KAPA HyperPlus Library Preparation Kit (KAPA Biosystems) according to the manufacturer’s instructions. During cDNA synthesis and library preparation we used 17 and 11 PCR cycles, respectively. Sequencing was performed on a NovaSeq 6000 (Illumina) with 25 million reads per sample.

Bioinformatics analysis

Reads were trimmed using cutadapt version 1.11 to remove the “QuantSEQ FWD” adaptor sequence. We checked the quality of the reads using the FastQC algorithm (Love et al., 2014). Count tables were generated using STAR (Dobin et al., 2013) (version 2.5.3) through mapping against the Genome Reference Consortium Human Build 38 patch release 10 (GRCh38.p10) combined with a general transfer format (GTF) file, both downloaded from the Ensembl database. The RNA-sequencing by Expectation Maximization (RSEM) (Li and Dewey, 2011) software (version 1.3.0) was used to produce the count tables.

Differential gene expression analysis was performed using R-studio (Version 1.1.456) with the edgeR (Robinson et al., 2009) and limma (Ritchie et al., 2015) packages. We included genes with a count per million greater than 1 in at least two samples. The trimmed mean of M values (Robinson and Oshlack, 2010) algorithm was used for normalization. The normalized counts were then transformed in a log2 fold-change (log2FC) table with their associated statistics. In each comparison, genes with a | log2FC | > 1 and an p-value < 0.05 were considered as significantly differentially expressed.

Gene set enrichment analysis (GSEA) was performed using the ‘investigate gene sets’ function on the website of the Molecular Signature Database (http://www.gsea-msigdb.org/gsea/msigdb/annotate.jsp). We used all differentially expressed genes (Supplementary Tables 2, 3) as input and investigated Hallmark, C2 and C5 gene sets with FDR-q values < 0.05. For the analysis of gene dosage effects, we could only use the TOP500 differentially expressed genes by p-value since the Molecular Signature Database changed the cut-off during the period of our data analysis (raw data, Supplementary Tables 26, 27). In the figures 1f and g we used a supervised approach and sorted the significantly deregulated gene sets by the total amount of genes in overlap. Figure 2f was also supervised but the gene sets were sorted by FDR-q value.

Human embryos with the highest number of genes with abnormal copy number show stronger p53 pathway and apoptosis response.

a. Supervised hierarchical clustering of the TOP 500 deregulated genes sorted by p-value. Color coding: blue = euploid, green = Low dosage and red = High dosage. b,c. Volcano plots after differential gene expression analysis with a cutoff value of |log2 fold change| > 1 and −log10(p-value) < 0.05 for low dosage versus euploid embryos (b) and high dosage versus euploid embryos (c). d,e, Venn diagrams comparing the upregulated (d) and downregulated (e) genes in low dosage versus high dosage embryos. f. Supervised plot of pathway analysis using the Hallmark library of the Molecular Signatures Database (FDR < 0.05) based on the list of the TOP 500 expressed genes by p-value between low or high dosage versus euploid embryos. Pathways are ranked based on FDR-q-values of both test sets, the most significant deregulated pathway ranked from top to least significant pathway. g,h. Dot plots of the genes in overlap with the Hallmark p53 pathway (g) and Hallmark apoptosis pathway (h) of both low and high dosage embryos sorted by log2 Fold Change from the highest to the lowest. Dots with an increased size and a darker blue represent a higher value of log2 Fold Change and lower p-value, respectively. i, Violin plots with box and whisker plots of the counts per million mapped reads of a supervised set of 10 genes (from Supplementary Table 2) that are part of apoptosis (BAD, BCL2L1, TNFRSF10B), p53 pathway (DRAM1, MDM2, CDKN1A/B, PURPL), growth arrest (GADD45A), and DNA-damage (DDB2). *p = 0.027, **p = 0.010, ***p = 0.004, ****p < 0.001 using the Jonkheere-Terpstra test. Box and whisker plots show median and minimum to maximum values.

Ingenuity Pathway analysis (IPA, Qiagen Inc.) was used to investigate activated, deregulated and inhibited upstream regulators and canonical pathways. As for the GSEA we used all differentially expressed genes of our two test sets with their respective log2FC and p-value. For the upstream regulators IPA computes z-scores with associated p-values that are considered significant when p < 0.05. If significant, a z-score > 2 means activated, a z-score < 2 means inhibited and in between means deregulated. For the canonical pathways IPA computes z-scores with associated p-values that are considered significant when p < 0.05. If significant, we considered a z-score > 1.5 as activated, a z-score < 1.5 as inhibited and in between as deregulated. We generated all figures in a supervised approach.

Transcription factor analysis of all differentially expressed genes was performed using Enrichr (https://maayanlab.cloud/Enrichr/). We selected the outputs from ChEA 2016 and ENCODE ChEA consensus and considered those transcription factors as significant with a p-value < 0.05.

The copy number variation (CNV) analysis was inferred from our RNA-sequencing data set using inferCNV R package (version 1.7.1) (Inferring copy number alterations from tumor single cell RNA-Seq data (https://github.com/broadinstitute/inferCNV/wiki”). Our count table was used as input and compared to the reference set which in our case was the expression data of the euploid embryos. The analysis followed the standard workflow in inferCNV with the following parameters: “denoise” mode, value of 1 was used for the “cutoff”, prediction of CNV states using default Hidden Markov Models (HMM) method, “cluster_by_groups” was set to TRUE, and other values were set by default.

All figures were generated using R-studio (Version 1.1.456).

Gene dosage analysis

We calculated the total number of loci with imbalanced gene expression based on the genetic abnormalities identified during the PGT. For this, we used the number of coding loci per chromosomal region as listed in the ENSMBL database (ensemble.org) and included both gains and losses, either mosaic or homogeneously present, and both full chromosomal aneuploidy as well as segmental abnormalities. The embryos were ranked based on the total number of imbalanced loci and separated into a low- and a high-imbalanced gene dosage group. The low-dosage group contained the embryos with the lowest 50th percentile in number of unbalanced loci, the high-dosage group the embryos with 50th percentile or higher (Supplementary Table 25, 29).

Immunocytochemistry

Embryos were fixed using 4% paraformaldehyde (Sigma-Aldrich) for 10 min, washed three times in PBS/2%BSA for 5 min and then permeabilized using 0.1% Triton X-100 (Sigma-Aldrich) for 20 min followed by another washing cycle. We used 10% FBS supplemented with 2% BSA for 1h to block non-specific protein binding. The embryos were then incubated overnight with primary antibodies (Supplementary Table 30) diluted in blocking buffer at 4°C. The next day, they underwent another washing cycle using PBS/2%BSA and were then incubated with secondary antibodies (1:200, Supplementary Table 26) for 1h at room temperature. After washing, nuclei were stained using Hoechst 33342 (5µg/mL, Life Technologies) diluted in PBS/2%BSA for 15 min. For imaging we mounted the embryos on poly-L-lysine coated glass slides (Sigma-Aldrich) in 3µL of PBS. To avoid flattening of the embryos we used Secure-SealTM Spacers (9mm diameter, 0.12mm deep, Thermofisher) before putting the coverslips in place. For the experiments that required re-staining (Fig. 4a) we imaged the embryos in 3µL PBS/2%BSA in 18 well µ-Slides (Ibidi) and subsequently recuperated them. After recuperation we photobleached the embryos stained for CASP3/7 by using maximum laser power of the 488 nm channel and maximum size of the pinhole for 10 min. The embryos were then re-stained using rat-anti-GATA4 primary antibody and the matching Alexa Fluor 488 secondary antibody (Supplementary Table 30) (to replace the CASP3/7 staining).

Imaging and image analysis

Confocal imaging was performed using the LSM800 (Zeiss). Z-stacks were taken with a LD C-Apochromat 40x/1.1 NA water immersion objective in optical sections of 1µm. Nuclei were counted using Arivis Vision 4D (3.4.0). All other measurements were performed using the Zen 2 (blue edition) software based on the optical sections and/or orthogonal projections.

Statistics

The type of statistical test and p-values are mentioned in the figure legends. For experiments that contained groups with small sample sizes we used non-parametric tests (Mann-Whitney or Kruskal-Wallis test). Fisher-exact test was used to determine the dependency of the ploidy status on the presence or absence of primitive endoderm and to test for differences in the quality of the ICM of euploid and aneuploid embryos. Otherwise we used Student’s t-test with Welch’s correction when we assumed normal distribution and unequal standard deviations. We used GraphPad Prism 9.0.0 or R-studio for statistical testing. The trend analysis regarding gene dosage effects was performed with SPSS using the Jonkheere-Terpstra test.

Results

Human embryos with meiotic complex aneuploidy show transcriptional signatures of a deregulated p53 pathway and apoptosis

We first tested the hypothesis that chromosomally unbalanced cells are cleared during human preimplantation development by apoptosis, mediated by both sustained p62-LC3B accumulation as a marker for autophagy and DNA-damage-mediated p53 activation. Fifty cryopreserved human blastocysts (5 or 6dpf) that had been previously diagnosed by Preimplantation Genetic Testing (PGT) as either euploid (n=28) or containing non-mosaic complex aneuploidy of at least two whole chromosomes and assumed of meiotic origin (except for one) (n=22, further referred to as aneuploid, Fig. 1a, Supplementary Table 1) underwent a second TE biopsy to collect cells for RNA sequencing. Eleven warmed human cleavage stage embryos that were treated with 0.5 µM reversine from 3dpf to 4dpf and further developed to the blastocyst stage (referred to as reversine-treated embryos, Fig. 1b) were also biopsied on 5dpf to serve as a comparison with the results obtained in the mouse. The concentration of 0.5 µM reversine was chosen based on previous experiments in mouse embryos (Bolton et al., 2016; Singla et al., 2020). Of these, 14 euploid, 20 aneuploid and 11 reversine-treated embryos (total=45) yielded good quality RNA-sequencing results.

Unsupervised hierarchical clustering and principal component analysis using all expressed genes revealed no clustering of the euploid, aneuploid or reversine-treated embryos in distinct groups (Fig. 1c, Supplementary Fig. 1a). We used InferCNV (https://github.com/broadinstitute/inferCNV/wiki”) to bioinformatically infer unbalanced chromosomal copy numbers from gene expression data, based on higher or lower expression of genes located on chromosomes with gains or losses respectively (Fig. 1d, Supplementary Fig. 1b, Supplementary Table 1). We compared the outcomes to the results obtained during PGT and found a match for 45/48 of the full chromosome aneuploidies, while euploidy was correctly predicted in all cases (Supplementary Fig. 1b). These results show that aneuploidy in human blastocysts results in significant abnormal gene dosage effects. The exact karyotypes of the reversine-treated embryos could not be determined since reversine induces different aneuploidies in each cell, leading to mosaic TE biopsies. However, InferCNV allowed us to determine a trisomy 12 meiotic chromosomal abnormality in one reversine-treated embryo (Supplementary Figure 1c), confirming that the reversine treatment causes random aneuploidy while naturally occurring meiotic aneuploidies are readily determined.

Despite the abnormal gene dosage, and in line with the lack of clear clustering in the PCA, the differential gene expression analysis yielded very few genes with a false discovery rate (FDR-q-value) under 0.05. These findings are similar to those of other studies on the transcriptome of aneuploid human embryos (Groff et al., 2019; Maxwell et al., 2022). Therefore, we used the differential gene expression with a cut-off value of |log2 fold change|>1 and −log10(p-value)<0.05, and primarily focused on pathway analysis. We found 547 upregulated and 463 downregulated genes in aneuploid versus euploid embryos (Fig. 1e, Supplementary Table 2). Amongst the top upregulated genes was Growth Differentiation Factor 15 (GDF15), a gene that has been found to be upregulated in human aneuploid cells (Dürrbaum et al., 2014) including aneuploid human embryos (Starostik et al., 2020). When comparing reversine-treated versus euploid embryos we observed 539 upregulated and 355 downregulated genes (Fig. 1f, Supplementary Table 3). While only 9.6% of the differentially expressed genes are in common between the aneuploid and reversine-treated embryos (Fig. 1g, h, Supplementary Table 4), pathway analysis revealed several commonly deregulated pathways. We identified deregulation of the p53 pathway and apoptosis (FDR q-value < 0.05, Fig. 1i, Supplementary Tables 5, 6) and deregulation of cell population proliferation, apoptotic process, regulation of cell death, and cell cycle (Fig. 1j, Supplementary Tables 7, 8). The prediction of p53 activation is also supported by differential expression of several p53 target genes (Jaiswal et al., 2021), such as MDM2 and PPM1D (feedback); CDKN1A, BTG2, GADD45A (cell-cycle arrest); TNFRSF10B (apoptosis), DDB2 and RRM2B (DNA-repair); FDXR (metabolism) and SESN1 (translation control) (Supplementary Table 2). Other pathways of interest related to DNA damage and p53, such as cellular response to DNA damage stimulus and DNA damage response signal transduction by p53 class mediator were specific to the aneuploid group. Cellular senescence signatures in the aneuploid embryos and selective autophagy in the reversine-treated group were also deregulated (Supplementary Tables 9, 10). Ingenuity pathway analysis (IPA) showed that upstream regulators CDK19 and AURK (cell cycle) and GAS2L3 and ANLN (cell growth) were inhibited in the aneuploid group (Fig. 1k, Supplementary Table 11). In line with the GSEA analysis, p53 was the most activated upstream regulator. Reversine-treated embryos activated ATM which is a well-known sensor of DNA-damage that mediates target genes such as p53 (Banin et al., 1998) (Fig. 1l, Supplementary Table 12). Canonical pathway analysis by IPA confirmed that p53 signalling, autophagy, cell-cycle pathways and senescence are deregulated in aneuploid embryos (Fig. 1m, Supplementary Table 13). The TE of reversine-treated embryos showed activated senescence, similar to the mouse reversine-treated embryos (Bolton et al., 2016) (Fig. 1n, Supplementary Table 14). Transcription factor analysis identified significant p53 expression in both aneuploid and reversine-treated groups (Supplementary Tables 15, 16).

We next compared our set of differentially expressed genes to those identified by other studies investigating the gene expression of euploid and aneuploid embryos, namely: mosaic monosomies 4, 7, 10, 15, 19 and mosaic trisomies 15 and 19 (Maxwell et al., 2022), trisomy 16 and monosomy 16 (Fuchs Weizman et al., 2019), monosomy 21 (Sanchez-Ribas et al., 2019) and embryos that contained various types of chromosomal abnormalities (Vera-Rodriguez et al., 2015; Licciardi et al., 2018; Starostik et al., 2020; Martin et al., 2023) (detailed overview in Supplementary Table 17). Overall, we found modest overlaps (Supplementary Figure 1d,e, Supplementary Tables 18-20). Most commonalities were with the data of Starostik et al, where we found an overlap in 87 genes (2.3%) (Starostik et al., 2020). These included target genes of the p53 pathway such as MDM2 and CDKN1A (Supplementary Fig. 1d, Supplementary Table 20). Interestingly, GSEA of these 87 genes identified an enrichment of intrinsic apoptotic signalling pathway by p53 class mediator and the apoptotic signalling pathway, in line with the findings of our dataset (Supplementary Table 19).

High gene-dosage imbalances result in stronger cellular stress signatures in human embryos

To test if there was a relationship between the level of stress signatures and the size of chromosomal imbalances in embryos as previously described in cancer (Dürrbaum and Storchová, 2016), we separated the aneuploid group into a low- and a high-imbalanced gene dosage group, based on the total number of coding loci that were affected by the chromosomal abnormalities. The low-dosage group contained the embryos under the 50th percentile when ranking the embryos from the lowest to the highest number of unbalanced loci, the high-dosage group the embryos with 50th percentile or higher (Supplementary Table 21). Hierarchical clustering using the top 500 deregulated genes showed that all high-dosage embryos clustered together and separated from the euploid embryos, while the low-dosage embryos clustered with either group (Fig. 2a). Differential gene expression with |log2 fold change|>1 and −log10(p-value)<0.05 resulted in 404 upregulated and 340 downregulated genes in low-dosage versus euploid embryos and in 716 upregulated and 627 downregulated genes in high-dosage versus euploid embryos with an overlap of 104 (10%) between both aneuploid groups (Fig. 2b-e, Supplementary Tables 22,23). We identified deregulation of TNFA signalling via NFkB, p53 pathway and apoptosis for both groups (FDR q-value < 0.05, Fig. 2f, Supplementary Tables 24,25). High-dosage embryos additionally showed an unfolded protein response, suggesting that this signature may only be present in the embryos with more and/or larger abnormal chromosomes. In the high-dosage group more of the genes of the p53 and apoptosis pathways were deregulated, and with higher significance (Fig. 2g,h, Supplementary Tables 26,27). The gene expression levels in the p53 and apoptosis pathways show a stepwise distribution between the gene dosage groups (Fig. 2i).

Aneuploidy triggers proteotoxic stress, autophagy and apoptosis in human embryos

To further investigate the cellular stress responses identified by the transcriptomic analysis, we immunostained the same embryos that underwent TE biopsy for transcriptome analysis. Thirty-seven (18 euploid, 19 aneuploid) of the 50 PGT embryos and 8 of the 11 reversine-treated embryos were successfully stained for markers of apoptosis, proteotoxic stress and autophagy (Fig. 3a,b). Aneuploid embryos showed increased intensity levels of active CASP3/7 (13 euploid and 8 aneuploid embryos stained) and CASP8 (6 euploid, 9 aneuploid and 8 reversine-treated embryos stained) (Fig. 3c-f). Both aneuploid and reversine-treated embryos had lower cell numbers than euploid embryos (Fig. 3g). We investigated whether proteotoxic stress and autophagy (Oromendia et al., 2012; Stingele et al., 2012; Dürrbaum et al., 2014; Ohashi et al., 2015; Singla et al., 2020) were also present in aneuploid human embryos. The number of puncta per cell of autophagy markers LC3B (19 euploid, 22 aneuploid and 8 reversine-treated embryos stained) and p62 (6 euploid and 9 aneuploid embryos stained) and the intensity levels of proteotoxic stress marker HSP70 (13 euploid, 8 aneuploid and 8 reversine-treated embryos stained) were significantly increased in both aneuploid and reversine-treated embryos (Fig. 3h-k). The gene expression profiles suggest that the accumulation of autophagic proteins in aneuploid embryos is caused by increased autophagic flux due to differential expression of the p53 target gene DNA Damage Regulated Autophagy Modulator-1 (DRAM1), rather than by inhibition of autophagy (Supplementary Table 2). DRAM1 is located in the lysosomal membrane and it increases autophagic flux and apoptosis after stress-induced p53 activation in cancer cells (Crighton et al., 2006; Guan et al., 2015).

Immunostaining reveals increased apoptosis, autophagy, proteotoxic stress and DNA-damage in chromosomally abnormal human embryos.

a,b. Same experimental set-up as described in Fig. 1 a,b, but focused on immunostaining analysis. c, Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white) and CASP3/7 (green). d, Orthogonal projections after immunostaining of euploid, aneuploid and reversine-treated embryos for DNA (white) and CASP8 (green). e, CASP3/7 mean intensity per cell. Euploid n=13, Aneuploid n=8 embryos. Mann-Whitney test, *p = 0.0199. f, CASP8 mean intensity per cell. Euploid n=6, Aneuploid n=9, Reversine-treated n=8 embryos. Kruskal-Wallis test, *p = 0.0394, **p = 0.0085. g, Number of nuclei per embryo. Euploid n=19, Aneuploid n=18, Reversine-treated n=8 embryos. Kruskal-Wallis test, ****p < 0.0001, *p = 0.0305. h, Orthogonal projections after immunostaining of euploid, aneuploid and reversine-treated embryos for DNA (white), LC3B (turquoise) and HSP70 (magenta). i, LC3B puncta per cell. Euploid n=19, Aneuploid n=22 embryos, Reversine-treated n=8 embryos. Kruskal-Wallis test, *p = 0.0189, **p = 0.0074. j, HSP70 mean intensity per cell. Euploid n=13, Aneuploid n=8, Reversine-treated n=8 embryos. Kruskal-Wallis test, **p = 0.0077, ***p = 0.0008. k, Orthogonal projections after immunostaining of euploid, aneuploid for DNA (white) and p62 (turquoise). l, p62 puncta per cell. Euploid n=6, Aneuploid n=9 embryos. Mann-Whitney test, **p = 0.0076.

Brightfield pictures were obtained during confocal imaging. All scale bars are 20 µm. Box and whisker plots show median and whiskers show minimum to maximum values. Bar plots show mean ± s.d. For all plots each dot represents a single embryo and ns = not significant.

Aneuploidy increases apoptosis in the trophectoderm and impairs the second lineage segregation in the inner cell mass

We further investigated if these effects are cell-type specific, as shown in the mouse (Bolton et al., 2016; Singla et al., 2020) and human 2D gastruloids (Yang et al., 2021), and how they affect the first lineage segregation event to TE and ICM, and the second lineage segregation event to EPI and PrE.

We co-stained a new batch of expanded blastocysts for OCT4 (10 euploid, 14 aneuploid, ICM/EPI) and CASP3/7 (10 euploid, 14 aneuploid), LC3B (4 euploid, 5 aneuploid) or Serine 15 p53 (3 euploid, 5 aneuploid) (Fig. 4a). To ensure that there was sufficient time for differentiation to EPI and PrE, we thawed blastocysts at 5dpf or 6dpf and cultured them for 16 hours. All information on embryo score and developmental day and stage at time of cryopreservation can be found in Supplementary Table 28. The latter marker was chosen to demonstrate that the transcriptomic signature of p53 signalling is caused by DNA-damage-induced phosphorylation of Serine-15 on p53 (Loughery et al., 2014). Aneuploid embryos showed a higher percentage of cells with activated p53 (Fig. 4b,c), as suggested by our transcriptomic data (Figure 1), although these differences did not reach significance when considering the ICM/EPI (OCT4-positive) cells or TE (OCT4-negative) separately (Supplementary Fig. 2a-c). We then investigated cell type specificity of proliferative defects and apoptosis. Both the aneuploid TE and OCT4-positive cells showed decreased cell numbers (Fig. 4d-f) that led to an overall decrease of cell numbers in whole aneuploid embryos (Fig. 4g), confirming our previous experiment (Fig. 3g). The observed increased percentage of nuclear CASP3/7 positive cells in the TE of aneuploid embryos (Fig. 4h,i) and increased CASP3/7 mean intensity per cell in whole aneuploid embryos (Fig. 4h,j) points to apoptosis as the cause of this decrease in cell numbers. However, no OCT4-positive cells of either euploid or aneuploid group showed nuclear CASP3/7 signal (Fig. 4h). We also found CASP3/7-positive micronuclei which were increased in number in aneuploid embryos (Fig. 4h, Supplementary Fig. 2d,e).

Aneuploid human embryos show less cells in trophectoderm and OCT4-positive cells and impaired second lineage segregation

a, 5 or 6dpf post-PGT-A euploid and fully aneuploid human blastocysts were warmed and kept for 16hrs in culture to allow for the second lineage segregation. The blastocysts were live-stained with CASP3/7 for 30min and fixed for immunostaining proteins of interest. In a first scan we analyzed OCT4 (ICM), CASP3/7 (apoptosis), LC3B (autophagy) and S15p53 (DNA damage) in both lineages. Embryos were then re-stained with GATA4 as a marker for PrE. b, Orthogonal projections after immunostaining of euploid, aneuploid for DNA (white) and Serine (S) 15 p53 (turquoise). c, Percentage (%) of Serine 15 p53 positive cells per embryo. Euploid n = 3, Aneuploid n = 5. Mann-Whitney test, *p = 0.0357. d, Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white) and OCT4 (magenta). The first aneuploid panel (Trisomy 14 and 16) shows a similar number of ICM cells compared to the euploid embryo. The second aneuploid panel (Trisomy 14 and 22) shows an embryo without an ICM. e,f,g, Differences in number of nuclei per embryo between euploid and aneuploid embryos in the OCT4-negative cells (trophectoderm) (e) **p = 0.0092, OCT4-positive cells (ICM/EPI), *p = 0.0412 (f) and whole embryo **p = 0.0067 (g). Euploid n = 10, Aneuploid n = 14. Student’s t-test with Welch’s correction. h, Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white), CASP3/7 (green) and OCT4 (magenta). i, Percentage (%) of CASP3/7 positive (+) cells in the trophectoderm lineage (OCT4-negative cells). Euploid n = 9, Aneuploid n = 14. Student’s t-test with Welch’s correction, *p = 0.0453. j, CASP3/7 mean intensity per cell of whole embryos. Euploid n = 9, Aneuploid n = 14. Student’s t-test with Welch’s correction, *p = 0.0332. k, Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white), OCT4 (magenta) and GATA4 (green). The first aneuploid panel (trisomy 13 and 15) shows an embryo with presence of a GATA4 positive cell (PrE). The second aneuploid panel (Trisomy 7 and monosomies 5, 12 and 21) shows an embryo that did not contain GATA4 positive cells. l, Differences in the number of cells per embryo that were OCT4 positive and either GATA4 negative (red) or positive (green) between euploid and aneuploid embryos. In case GATA4-positive cells (PrE) were present we considered the GATA4-negative cells to be epiblast (EPI). All euploid embryos contained GATA4-positive cells, 7/14 aneuploid embryos had an ICM completely lacking GATA4-positive cells (Fisher-exact test, p = 0.0188). m, Differences in the PrE/EPI ratio between euploid and aneuploid embryos that had an ICM containing GATA4-positive cells. Euploid n = 9, Aneuploid n = 7. Note here that one euploid embryo (#6, l) was lost during re-staining for GATA4. Mann-Whitney test, **p = 0.0089. n, Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white), LC3B (turquoise) and OCT4 (magenta). Yellow circle indicates the ICM. The first aneuploid panel (Trisomy 1 and 20 and monosomy 18) shows OCT4-positive cells (ICM/EPI) with low levels of autophagy and, after re-staining for GATA4, without PrE (right panel, ICM zoom). The second aneuploid panel (Monosomy 14) shows an embryo with high levels of autophagy in the ICM that contains PrE cells (right panel, ICM zoom). o,p,q. Differences in LC3B puncta per cell between euploid and aneuploid embryos in the whole embryo,*p = 0.0317 (o),(OCT4-negative cells (trophectoderm),*p = 0.0317 (p) and OCT4-positive cells (ICM/EPI) (q). Euploid n = 4, Aneuploid n = 5. Mann-Whitney test. r, Violin plots with box plots of OCT4-positive-cells (ICM/EPI) per embryo (left, ns), OCT4-negative cells (trophectoderm) per embryo (center, *p = 0.043) and GATA4-positive-cells (PrE) per embryo (right, ****p < 0.001, Jonkheere-Terpstra test) depending on gene dosage. Euploid (light blue), Low dosage (middle blue) and High dosage (darkest blue).

Yellow arrows indicate presence of the signal and yellow arrow heads indicate absence of signal. Brightfield pictures were obtained during confocal imaging. All scale bars are 20 µm. Box and whisker plots show median and minimum to maximum values. Bar plots show mean ± s.d. For all plots each dot represents a single embryo; ns = not significant. T = Trisomy, M = Monosomy

The observation (figure 4f) that the ICM/EPI cell numbers of aneuploid embryos fell into two groups with either low or high cell counts led us to further investigate the impact of aneuploidies on the second lineage segregation event. All blastocysts were re-stained for GATA4 (PrE marker) and imaged together with OCT4 (ICM/EPI marker). We found that while all euploid embryos contained GATA4-positive cells, 7/14 aneuploid embryos had no GATA4-positive cells (Fig. 4k,l) and that in those with GATA4-positive PrE cells, there was a significantly lower ratio of GATA4-positive to OCT4-positive/GATA4-negative cells (Fig. 4m). In line with their decreased ICM cell numbers and absence of PrE, the aneuploid embryos had received a significantly lower ICM score by the clinical embryologists at time of cryopreservation (1xA/13xB) when compared to euploid embryos (6xA/4xB, p=0.0088, Fisher-exact test).

We then investigated whether there are different levels of LC3B as marker of autophagy between the aneuploid TE and ICM in an additional 5 aneuploid and 4 euploid embryos. Aneuploid embryos displayed increased levels of autophagy mainly attributable to the TE (Fig. 4n-q), while their ICM/EPI cells showed a small increase in autophagy which did not reach significance (Fig. 4q). The aneuploid embryos could be separated in two groups: a first group with low levels of autophagy lacking GATA4-positive cells in 2 out of 3 embryos (Fig. 4n, left and right panels, 2nd row) and a high-level autophagy group in which both embryos contained GATA4-positive cells (Fig. 4n, left and right panels 3rd row).

Lastly, we separated the aneuploid embryos into a low- and a high-dosage group as done for the gene-expression analysis (Supplementary Table 29). Trend analysis showed that the higher the number of imbalanced loci, the lower the cell numbers in the TE and PrE but not in the EPI (Fig. 4r).

Discussion

In this work, we studied the consequences of aneuploidy in human preimplantation embryos that carried a uniformly present, presumed meiotic, complex aneuploidy. In our study we find that, while aneuploid embryos had lower cell numbers in both TE and ICM/EPI, the effect of aneuploidy was cell-type dependent. Aneuploid TE presented transcriptomic signatures of p53 signalling and apoptosis, and immunostaining of whole embryos showed that aneuploidy results in increased proteotoxic stress, autophagy, DNA damage-mediated activated p53 and subsequent apoptosis in the TE. In contrast, aneuploidy impairs the second lineage differentiation and affects PrE formation in particular.

Overall, our gene-expression results are in line with previously published RNA-sequencing studies on aneuploid human embryonic cells that identified transcriptional signatures consistent with DNA damage (Vera-Rodriguez et al., 2015; Licciardi et al., 2018; Fuchs Weizman et al., 2019), alteration of cell proliferation (Starostik et al., 2020; Maxwell et al., 2022; Martin et al., 2023), the cell cycle (Vera-Rodriguez et al., 2015; Licciardi et al., 2018; Fuchs Weizman et al., 2019; Starostik et al., 2020; Martin et al., 2023), deregulation of autophagy (Licciardi et al., 2018; Sanchez-Ribas et al., 2019), p53 signalling (Licciardi et al., 2018) and apoptosis (Licciardi et al., 2018; Groff et al., 2019; Sanchez-Ribas et al., 2019; Maxwell et al., 2022; Martin et al., 2023). However, when comparing our dataset of differentially expressed genes to these studies, we only found modest overlaps at the level of individual genes. This is probably because embryos of different developmental stages were included in the studies, containing specific aneuploidies, and of different sample types (overview in Supplementary Table 17). Naturally acquired and induced aneuploidies only share a few differentially expressed genes but show overall equal signatures after pathway analysis, suggesting that reversine-based models of aneuploidy are very close but may not fully mimic endogenous aneuploidy. We also observed aneuploidy-mediated gene dosage effects in the TE cells, as extensively reported in human and mouse aneuploid cancer cell lines (Dürrbaum and Storchová, 2016), and found that a higher dosage leads to stronger p53 signalling and apoptosis responses.

We next tested the predictions made based on the gene-expression profiles with immunostaining. First, we found evidence of proteotoxic stress and increased expression of autophagy proteins, as well as activation of apoptosis. The most likely series of events is that p62/LC3B/CASP8 interaction results in autophagy-mediated apoptosis in aneuploid human embryos. Nevertheless, we cannot rule out indirect activation of apoptosis mediated by autophagy due to the degradation of cell organelles (Gump and Thorburn, 2011). This is supported by the observation in the aneuploid mouse embryo model that autophagy is independent of the mTOR pathway (Singla et al., 2020). Other upstream regulators of autophagy could be involved, such as the cGAS-STING pathway (Krivega et al., 2021) or, as also observed in our dataset, p53-mediated activation of DRAM1 (Crighton et al., 2006) that can induce apoptosis by increasing BAX protein levels (Guan et al., 2015). Second, we confirmed the presence of activated p53 in aneuploid human embryos. We specifically stained for p53 with phosphorylation on Serine 15, mediated by kinases such as ATM and ATR as a direct result of DNA-damage (Loughery et al., 2014). This indicates that aneuploidy also induces the DNA-damage response in the embryo, as found in the mouse and 2D gastruloids (Singla et al., 2020; Yang et al., 2021). An interesting observation is that aneuploid embryos contained increased numbers of CASP3/7 positive micronuclei. This is similar to findings in a non-human primate model showing that abnormal chromosomes can be encapsulated in micronuclei, which can subsequently be selectively eliminated from the embryo (Daughtry et al., 2019). In human cleavage stage embryos these micronuclei have been shown to contain genetic material that originates from chromosome breakages due to replication fork stalling and DNA-damage (Palmerola et al., 2022), further highlighting how aneuploidy is leading to DNA-damage in the human embryo. In sum, our results in this part are in line with previous findings on mouse and 2D gastruloid models (Singla et al., 2020; Yang et al., 2021), and suggest that these conserved stress responses to aneuploidy are also activated in fully aneuploid embryos with complex karyotypes, both converging on apoptosis if not resolved. First, proteotoxic stress resulting from gene-dosage imbalances mediates the unfolded protein response and autophagy, and second, replication stress that can lead to DNA-damage, resulting in p53 activation and ultimately to apoptosis (Li et al., 2010; Thompson and Compton, 2010; Santaguida et al., 2015, 2017; Zhu et al., 2018; Chunduri and Storchová, 2019; Krivega et al., 2022; Regin et al., 2022).

In the second part of our work, we focused on studying the lineage-specific response to complex aneuploidy, as the studies on reversine-treated mouse embryos (Bolton et al., 2016; Singla et al., 2020) and human 2D gastruloids (Yang et al., 2021) also identified a cell-type specific effect of aneuploidy. In the mouse, aneuploid cells are eliminated by apoptosis in the EPI, while tolerated in the TE, and in human 2D gastruloids, aneuploid cells are eliminated in the post-gastrulation embryonic germ layers but tolerated in the TE. We found lower cell numbers and increased apoptosis in aneuploid TE cells. While we also detected lower cell numbers in aneuploid OCT4-positive ICM/EPI, we did not find any of them to be CASP3/7 positive, i.e. undergoing apoptosis. These results are in line with the findings of Victor et al. (Victor et al., 2019), where the TE of mosaic and aneuploid embryos showed increased levels of apoptosis, and the recent RNA-sequencing of TE and ICM samples of mosaic human embryos (Martin et al., 2023). Mosaic TE cells showed disrupted regulation of apoptosis while the ICM showed primarily deregulated mitochondrial function. However, in contrast to our findings, some of the embryos in the study of Victor et al. showed CASP3/7 positive signals in the ICM; further research will be needed to understand if these differences with our findings and those of Martin et al. are related to technical or to biological factors. Finally, the differences with the mouse results are possibly due to species-specific properties of the ICM/EPI cells. For instance, the lack of apoptosis in the human ICM/EPI could be due to an uncoupling between aneuploidy-induced stress signals and apoptosis, as observed in human embryonic stem cells (Mantel et al., 2007). In this context, the lower cell numbers in aneuploid embryos could be explained by two other known effects of p53 activation which are not apoptosis. First, p53 activation could decrease cell proliferation, leading to lower cell numbers. Secondly, p53 activation may downregulate OCT4 and NANOG expression, as seen in mouse embryonic stem cells (Lin et al., 2005), leading to loss of pluripotency and driving the cells to differentiation and out of the ICM/EPI.

Another difference is that we found that the human aneuploid TE had increased levels of autophagy markers, which did not reach statistical significance in the ICM/EPI, while the mouse EPI does show increased autophagy (Singla et al., 2020). Interestingly, the aneuploid OCT4-positive cells separated into a high and a low-level autophagy group, where the low-level group seemed to lack GATA4-positive PrE cells. Although based on a few samples, this data suggests that an ICM in a transcriptionally earlier state of pluripotency, i.e. GATA4-negative and prior to the second lineage segregation event, might be less sensitive to the consequences of aneuploidy, i.e. show less autophagy. We thus propose that aneuploid human embryonic cells in the ICM become more sensitive from the second lineage segregation onwards. We hypothesize that this effect may even become stronger and more specific to the embryonic lineage after onset of gastrulation, as elegantly shown in human 2D gastruloids (Yang et al., 2021).

Finally, we found that in those embryos that had already undergone the second lineage segregation event, the aneuploid embryos had a significantly lower ratio of PrE (GATA4-positive) to EPI (OCT4-positive/GATA4-negative) cells than euploid embryos. Our findings are in line with the upregulation of the PrE marker SOX17 that has been recently found by gene-expression analysis of euploid and low- and high level mosaic human embryos (Martin et al., 2023). Further, we observed that this effect was dosage-dependent, with embryos with larger gene-dosage imbalances showing significantly lower cell numbers in the TE and PrE than euploid embryos or embryos with lower gene-dosage imbalances. This suggests that complex aneuploidy both affects TE and ICM proliferation and PrE formation or that the PrE is more sensitive to aneuploidy-mediated apoptosis than the less differentiated EPI. An alternative explanation is that in aneuploid human embryos the PrE forms at a later time point due to their delayed development. In our study, to minimize the risk of missing the second lineage segregation, we purposefully used 5-6dpf expanded blastocysts that were kept in culture for an additional 16 hours after warming, to allow the embryos sufficient time to recover and develop.

When interpreting the results of this study, it is important to bear in mind that we identified the ICM/EPI as staining positive for OCT4 and the PrE as positive for GATA4. We confirmed that the embryos analysed were expanded blastocysts which had undergone the first lineage segregations, as OCT4 was restricted to the inner cells and absent in TE cells (Niakan and Eggan, 2013). The identity of human ICM and EPI cells is currently under vigorous debate (Meistermann et al., 2021; Radley et al., 2022) and exclusive markers of either cell type are untested. Therefore we cannot assign ICM or EPI identity to OCT4-positive cells. However, in blastocysts with GATA4-positive cells the second lineage segregation had occurred and PrE was established, and consequently we confidently assigned the EPI identity to OCT4-positive cells in these blastocysts. In GATA4-negative embryos OCT4-positive cells could be either ICM or EPI cells. Interestingly, we never found GATA4-positive cells in the absence of OCT4-positive cells, supporting the hypothesis that PrE cells differentiate from EPI cells (Meistermann et al., 2021). Limitations of our study are that we did not analyse the transcriptome of the ICM of our embryos, because after the biopsy they were used for immunostainings instead, and we can therefore not draw any conclusions on their transcriptome. We did not study embryos with the same type of aneuploidy and can thus not provide insight on chromosome-specific effects. Although some of our embryos carried genetic abnormalities in mosaic form as part of their complex karyotype (aneuploid/aneuploid mosaics), we did not study diploid/aneuploid mosaic embryos and can therefore not assert that these also activate the same cellular stress pathways in their aneuploid cells. Since these pathways are common to most cell types, our guess would however be that they are activated in mosaic embryos too. Lastly, the possibilities for functional studies and lineage tracing experiments in human embryos are very limited, which is why we can only present an observational study.

Taken together, our results suggest that while the same pathways appear to be activated in human and mouse embryos with complex aneuploidy, human embryos have a different response when it comes to the downstream effect of these pathways in the different cell lineages. While the mouse readily depletes aneuploid cells by apoptosis specifically from the embryonic lineage before implantation (Bolton et al., 2016; Singla et al., 2020), in the preimplantation human embryo, apoptosis appears to be more active in the TE, as also seen by others (Victor et al., 2019; Martin et al., 2023). The results suggest that in the human embryo, undifferentiated pluripotent cells have a different response to aneuploidy-induced cellular stress than the neighbouring differentiated cells.

Our results on human embryos with complex meiotic aneuploidy provide further insight into the mechanisms by which human embryos respond to gene-dosage imbalances, and may contribute to understanding how aneuploid cells are selectively eliminated in mosaic embryos. Further research, especially by using human embryos with endogenous chromosomal mosaicism, will be needed to shed light on the interactions between proteotoxic stress, the p53 pathway and differentiation during early development, for which the most novel models of human embryo implantation (Deglincerti et al., 2016; Shahbazi et al., 2016, 2020; Kagawa et al., 2021) will prove to be of incalculable value.

Supporting information

eLife Assessment

This study has uncovered some important initial findings about cellular responses to aneuploidy through analysis of gene expression in a set of donated human embryos. While the study's findings are in general solid, some experiments lack statistical power due to small sample sizes. The authors should try to get much more insight with their data highlighting the novel findings.

Reviewer #1 (Public Review):

This study investigated an important question in human reproduction: why most fully aneuploid embryos is incompatible with normal fetal development. Specifically, the authors investigated the cellular responses to aneuploidy through analysis of gene expression in a set of donated human blastocysts. The samples included uniform aneuploid embryos of meiotic origin and mosaic aneuploid embryos from the SAC inhibitor reversine treatment. The authors relied mainly on low-input RNA sequencing and immunofluorescence staining. Pathway analysis with RNA-seq data of trophectoderm cells suggested activation of p53 and possibly apoptosis, and this cellular signature appeared to be stronger in TE cells with a higher degree of aneuploidy. Immunostaining also found some evidence of apoptosis, increased expression of HSP70 and autophagy in some aneuploid cells. With combinational OCT4 and GATA4 as lineage markers, it appeared that aneuploidy could alter the second lineage segregation and primitive endoderm formation in particular.

Although this study is largely descriptive, it generated valuable RNA-seq data from a set of aneuploid TE cells with known karyotypes. Immunostaining results in general were consistent with findings in mouse embryos and human gastruloids.

While there is a scarcity of human embryo materials for research, the lack of single cell level data limits further extension of the presented data on the consequences of mosaic embryos. A major concern is that the gene list used for pathway analysis is not FDR controlled. It is also unclear how the many plots generated with the "supervised approach" were actually performed. The authors also appear to have ignored the possibility that high-dosage group could have a higher mitotic defects. Assuming a fully aneuploid embryo, why do only some cells display p53 and autophagy marker? The conclusion about proteotoxic stress was largely based on staining of HSP70. It appears from Figure 3 d,h that the same cells exhibited increased HSP70 and CASP8 staining. Since HSP70 is known to have anti-apoptotic effect, could the increased expression of Hsp70 be an anti-apoptotic response?

Reviewer #2 (Public Review):

A high fraction of cells in early embryos carry aneuploid karyotypes, yet even chromosomally mosaic human blastocysts can implant and lead to healthy newborns with diploid karyotypes. Previous studies in other models have shown that genotoxic and proteotoxic stresses arising from aneuploidy lead to the activation of the p53 pathway and autophagy, which helps eliminate cells with aberrant karyotypes. These observations have been here evaluated and confirmed in human blastocysts. The study also demonstrates that the second lineage and formation of primitive endoderm are particularly impaired by aneuploidy.

This is a timely and potentially important study. Aneuploidy is common in early embryos and has a negative impact on their development, but the reasons behind this are poorly understood. Furthermore, how mosaic aneuploid embryos with a fraction of euploidy greater than 50 % can undergo healthy development remains a mystery. Most of our current information comes from studies on murine embryos, making a substantial study on human embryos of great importance. However, there are only very few new findings or insights provided by this study. Some of the previous findings were reproduced, but it is difficult to say whether this is a real finding, or whether it is a consequence of a low sample number. The authors could get much more insight with their data.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This study has uncovered some important initial findings about cellular responses to aneuploidy through analysis of gene expression in a set of donated human embryos. While the study's findings are in general solid, some experiments lack statistical power due to small sample sizes. The authors should try to get much more insight with their data highlighting the novel findings.

We thank the editor for considering our manuscript for publication at elife, and for the helpful and thorough reviews of our work. Based on the suggestions of the reviewers, we have carried out additional experiments, expanded the sample size and reanalyzed the data. This has resulted in a thoroughly revised manuscript and much improved work, which we are convinced meets the requirements to be published as a version of record. Of note, the experiments for the revision required the support by 2 additional researchers from our lab which are now coauthors.

These are the main changes made to the initial manuscript:

(1) The RNA-seq data (Figures 1+2) is now FDR corrected and been reanalyzed. This has not affected the initial observations on the activation of p53 and apoptosis in aneuploid human embryos, as well as that the transcriptomic changes are driven by gene dosage effects.

(2) We have included the transcriptome analysis of reversine-treated embryos in the supplementary data.

(3) For validation of novel findings such as the presence of DNA-damage and the expression of DRAM1 in aneuploid embryos, we now include the stainings of 30 human blastocysts (Figure 3o-t). We found absence of DNA-damage in aneuploid embryos and that DRAM1 is increased in the TE but not the ICM of aneuploid embryos.

(4) We re-analyzed the co-expression of CASP8/HSP70 in reversine-embryos as suggested by reviewer 1 and found that both proteins tend to be co-expressed.

(5) We have added a new analysis of NANOG expression (Figure 4a,b) of the embryos used in Figure 3o-t and have found retention of NANOG protein in both the TE and ICM.

(6) We have added 6 euploid and 4 aneuploid embryos to Figure 4l-s, which support the conclusions on the absence of autophagy activation in the ICM and failure of PrE formation in aneuploid embryos.

(7) We have significantly changed the layout of the figures, revised the supplementary tables, added source data files and rewritten the discussion.

Regarding the sample size of the study, it is important to emphasize that human embryos are ethically sensitive material and that those with the specific genetic content we used in this study are rare, limiting our ability to expand the sample size. For the revision, we have added 40 human blastocysts to our initial 85 embryos. Compared to similar and high-quality studies using human embryos, our study shows a relatively large sample size (n=125): Victor et al. 2021: 30 human blastocysts for immunostainings1; Martin et al. 2023: 14 human blastocysts2; Martin et al. 2024: 64 human blastocysts3; Domingo-Muelas et al. 2023: 23 human blastocysts4.

Public Reviews:

Reviewer#1(PublicReview):

This study investigated an important question in human reproduction: why most fully aneuploid embryos is incompatible with normal fetal development. Specifically, the authors investigated the cellular responses to aneuploidy through analysis of gene expression in a set of donated human blastocysts. The samples included uniform aneuploid embryos of meiotic origin and mosaic aneuploid embryos from the SAC inhibitor reversine treatment. The authors relied mainly on low-input RNA sequencing and immunofluorescence staining. Pathway analysis with RNA-seq data of trophectoderm cells suggested activation of p53 and possibly apoptosis, and this cellular signature appeared to be stronger in TE cells with a higher degree of aneuploidy. Immunostaining also found some evidence of apoptosis, increased expression of HSP70 and autophagy in some aneuploid cells. With combinational OCT4 and GATA4 as lineage markers, it appeared that aneuploidy could alter the second lineage segregation and primitive endoderm formation in particular.

Although this study is largely descriptive, it generated valuable RNA-seq data from a set of aneuploid TE cells with known karyotypes. Immunostaining results in general were consistent with findings in mouse embryos and human gastruloids.

We thank the reviewer for the thorough evaluation of our manuscript. We have implemented most of the suggestions, which have further strengthened the original findings.

While there is a scarcity of human embryo materials for research, the lack of single cell level data limits further extension of the presented data on the consequences of mosaic embryos.

We did not include single cell RNA-seq data of mosaic human embryos in our study because we focused on embryos diagnosed with complex meiotic abnormalities. Our hypothesis was that the cellular consequences of aneuploidy would be strongest in this type of aneuploidies and most evident to identify and would allow us to provide a basis for the mechanisms of elimination of aneuploid cells in human embryos. In the manuscript (lines 596-626) we acknowledge the limitations of the extrapolation of our results to mosaic embryos.

A major concern is that the gene list used for pathway analysis is not FDR controlled. It is also unclear how the many plots generated with the "supervised approach" were actually performed.

We agree with the concerns about the fact that our differential expression gene list was not FDR but p-value ranked. We followed the suggestion of the reviewer and revised the RNAseq analysis and focused primarily on pathway analysis. We have also added the comparison between aneuploid and reversine treated embryos to the supplementary data and expanded the analysis of high dosage and low dosage embryos. Importantly, the new analysis has not changed the original finding that aneuploid embryos show hallmarks of p53 activation and apoptosis, and that these effects are gene dosage dependent. The manuscript now includes two completely revised and new figures 1 and 2.

Since we discarded the data generated from our previous approach, we do not use the term supervised approach anymore.

The authors also appear to have ignored the possibility that high-dosage group could have a higher mitotic defect.

This is indeed a possibility. In the discussion (lines 504-508) we have now incorporated the notion that the high dosage embryos could have higher mitotic defects, although our data cannot provide any evidence for this. Of note, the gene expression data shows that all aneuploid embryos (including low dosage and reversine embryos) equally show an enrichment for mitotic spindle pathway genes.

Assuming a fully aneuploid embryo, why do only some cells display p53 and autophagy marker?

This is a very good question, on which we can only speculate, but the answer likely lies in the diversity across cells of the same embryo.

Even in genetically homogenous tissues and cell cultures, individual cells can exhibit different levels of stress responses, such as p53 activation and apoptosis. This variation may be influenced by the local cellular environment, stochastic gene expression, or differences in cell cycle stages. Other studies on fully aneuploid human embryos could also not detect apoptotic responses in every cell1,3.

For instance, p53 activation differs even between cells that have a similar number of DNA breaks, and this activation is influenced by both cell-intrinsic factors and previous exposure to DNA damage5.

Cell cycle tightly regulates the response of cells to different stressors. For instance, cells in G1 or S-phase might be more sensitive to apoptosis signals6, while those in G2/M might escape this response temporarily7. Autophagy is more induced in G1 and S phases, with reduced activity in G2 and M phases8.

Individual cells may also have different levels of success in the activation of the compensatory pathways, including the unfolded protein response, autophagy, or changes in metabolism, resulting in some cells adapting better than others.

The expression of p53 and the sensitivity to apoptosis could also be influenced by epigenetic differences between cells, which may alter their transcriptional response to aneuploidy. Even in a genetically identical population, cells can have different epigenetic landscapes, leading to heterogeneous gene expression patterns.

The conclusion about proteotoxic stress was largely based on staining of HSP70. It appears from Figure 3 d,h that the same cells exhibited increased HSP70 and CASP8 staining. Since HSP70 is known to have anti-apoptotic effect, could the increased expression of Hsp70 be an anti-apoptotic response?

Our conclusion about proteotoxic stress was not solely based on HSP70 expression. We also stained for LC3B and p62, which are markers for autophagy and when highly expressed indirectly point towards underlying proteotoxic stress in the cells.

We reanalyzed the imaging of the stainings in the reversine-treated embryos, and found that the same cells were positive for both HSP70 and CASP8 staining while the minority was single positive (shown now in Figure 3k,l).

HSP70 does indeed not only unfold misfolded and aggregated proteins but does also have a function during cell survival and apoptosis9. HSP70 has been for instance found to inhibit the cleavage of Bid through active CASP8 within the extrinsic apoptosis pathway10. It is thus possible that it temporarily plays this role, and we have acknowledged this in the discussion (lines 623-626). On the other hand, the evidence points at an active apoptosis in the TE, with concomitant cell loss, so if HSP70 is indeed having an anti-apoptotic effect, it is having a limited impact.

Reviewer #2 (Public Review):

A high fraction of cells in early embryos carry aneuploid karyotypes, yet even chromosomally mosaic human blastocysts can implant and lead to healthy newborns with diploid karyotypes. Previous studies in other models have shown that genotoxic and proteotoxic stresses arising from aneuploidy lead to the activation of the p53 pathway and autophagy, which helps eliminate cells with aberrant karyotypes. These observations have been here evaluated and confirmed in human blastocysts. The study also demonstrates that the second lineage and formation of primitive endoderm are particularly impaired by aneuploidy.

This is a timely and potentially important study. Aneuploidy is common in early embryos and has a negative impact on their development, but the reasons behind this are poorly understood. Furthermore, how mosaic aneuploid embryos with a fraction of euploidy greater than 50 % can undergo healthy development remains a mystery. Most of our current information comes from studies on murine embryos, making a substantial study on human embryos of great importance. However, there are only very few new findings or insights provided by this study. Some of the previous findings were reproduced, but it is difficult to say whether this is a real finding, or whether it is a consequence of a low sample number. The authors could get much more insight with their data.

We thank the reviewer for the thorough evaluation of our manuscript and the valuable suggestions made in the private recommendations. We have expanded the sample size and have carried out additional experiments that have significantly improved the manuscript.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

(1) Instead of using cut off to generate a list, the authors could just rank the entire detected transcriptome for GSEA. This method fits better the authors' intentions of "primarily focused on pathway analysis." The cut-off value "-log10(p-value)<0.05" is not correct. As we can see from the PCA plot, one would not expect many cut off defined DEGs at all. The most obvious transcriptome change is dosage dependent, as the authors cleared showed with InferCNV.

We thank the reviewer for this suggestion and agree that this was an important concern of the study. We have entirely revised the RNA-seq analysis based on the proposed approach (Figure 1 and 2, Supplementary Figure 1). Also, we have included the analysis of aneuploid versus reversine treated embryos, which has allowed us to determine the differences between naturally occurring chromosomal abnormalities and those that are induced using reversine (Supplementary Figure 1).

We first performed differential gene expression analysis using DESEq2 with a cut-off value for significantly differentially expressed genes of | log2FC | > 1 and an FDR < 0.05. Based on the PCAs and the low number of differentially expressed genes for all comparisons, besides high dosage versus euploid embryos, we focussed primarily on pathway analysis.

For that, based on the reviewer’s suggestion, we generated a ranked gene list using the GSEA software (version 4.2.2, MSigDatabase) based on the normalized count matrix of the whole transcriptome that was detected after differential gene expression. The ranked gene list was then subjected to the run GSEA function, and we searched the Hallmark and C2 library for significantly enriched pathways. Thus, we could generate normalized enrichment scores, allowing us to predict whether a pathway is activated or suppressed. The details of the new analysis are described in the Material and Methods section (lines 220-232). Significance was determined using a cut-off value of 25% FDR. This cut-off is proposed in the user guide of the GSEA (https://www.gsea-msigdb.org/gsea/doc/GSEAUserGuideTEXT.htm) especially for incoherent gene expression datasets, as suggested by our PCAs, which allows for hypothesis driven validation of the dataset.

Indeed, we found that the most important transcriptome changes are aneuploidy dosage dependent. High dosage embryos show signatures of cellular unfitness, while low-dosage embryos still seem to activate survival pathways (lines 349-364).

This new analysis did not only increase robustness of our results but also introduced novel findings, which pave the road for future studies.

The validity of our findings is supported by recent work by the Zernicka-Goetz lab. We found that hypoxia is upregulated in low dosage human aneuploid TE cells. In line with our data, the Zernicka-Goetz lab found in a mouse model of low degree chromosomal abnormalities that hypoxia inducible factor 1A (HIF1A) promotes survival of extraembryonic aneuploid cells by reducing levels of DNA damage11.

(2) It would be very helpful if the authors could perform co-staining of multiple stress markers to better understand the origins of apoptosis and autophagy cells. In Fig 3d and 3h, it seems that the same reversine treated embryo was stained with CASP8, LC3B and HSP70. Is there any correlation between CASP8 and HSP70 at the single cell level? Is there any correlation between p53 and LC3B as the authors suggested, possibly through DRAM1?

We decided to use the complex aneuploid embryos that were left at our facility for the validation of novel findings such as upregulation of DRAM1 and presence and consequences of DNA damage in aneuploid embryos. As suggested by the editor and the other reviewer we also added embryos to existing datasets to increase the sample size where necessary. Therefore, we did not include other co-staining’s of multiple stress markers.

Following the reviewer’s suggestion, we reanalyzed the existing stainings and evaluated whether there is a correlation between CASP8 and HSP70 at the single cell level. The reversine-treated embryos were the only embryo group that was co-stained for both CASP8 and HSP70. We quantified the percentage of cells that were single or double positive for CASP8 and HSP70 and found a higher proportion of double positive cells than to single positives. Therefore, we concluded that there is indeed a correlation between both proteins at the single cell level in reversine-treated embryos and included this data in Figure 3k,l.

During the experiments for the revision, we found that the DRAM1 protein was upregulated in the cytoplasm of TE cells but not in the ICM of aneuploid embryos (Figure 3s,t), which validates the findings of the gene expression analysis. This data also supports our findings that autophagy is active in aneuploid TE cells while not significantly increased in aneuploid pluripotent ICM cells. Unfortunately, we could not stain LC3B and DRAM1 in the same embryo because the antibodies were raised in the same species.

(3) While " the possibilities for functional studies and lineage tracing experiments in human embryos are very limited," the authors can leverage in silico modelling (ie, PMID: 28700688) to address the roles of aneuploidy in blastocyst formation and development. Is there any selfregulating mechanism underlying the ratios of PrE and EPI? Is apoptosis of ICM cells a natural process during PrE formation (PMID: 18725515)?

It is a very interesting proposal to use in silico modelling to address the roles of aneuploidy during human blastocyst formation and lineage segregation. Although this type of analysis would yield very important insights, we are not able to address this point of the revision due to lack of expertise for this type of analysis in our group, requiring setting up a collaboration with experts in this field. In the discussion we proposed that future studies can leverage our data to be carried out in silico modelling and cited the proposed article (lines 608-610).

On the second part of the question, we would like to discuss the differences between mouse and human embryo studies. Parts of this were included in the discussion on the possible mechanisms of PrE elimination.

Is there a self-regulating mechanism for EPI/PrE formation?

To extrapolate the knowledge on mouse development to human it is important to bear in mind that (1) human embryos are outbred, as compared to inbred super-fertile laboratory mouse strains and (2) the embryos are donated to research by subfertile couples, which could compromise the EPI/PrE ratios. For instance, Chousal and colleagues found that poor quality blastocysts have a reduced number of PrE cells12. In human embryos the proportion EPI and PrE cells is indeed highly variable (20%-60%) and while the number of EPI cells does not increase between dpf6 and 7, the number of PrE cells does grow13. We found a similar variable number of EPI and PrE in our study on the lineage segregation mechanisms in good quality human embryos, with an absolute number of EPI of 12.1±6.5 cells and 8.4±3.44 PrE cells14.

By comparison, in late mouse blastocysts, the ratio EPI/PrE cells is consistent (2/3)15. Overall, self-regulating mechanisms in the human embryo are not yet studied in detail due to the lack of possible functional testing.

Is apoptosis a natural process during PrE formation?

Yes, in mice apoptosis is a natural process during PrE formation to eliminate misallocated cells of the inner cell mass through cell competition16,17. Yet, in the human embryo there is no evidence of such mechanisms. Although apoptosis is present even in human blastocysts of good quality18, the origin of such apoptotic cells is now still shown, although suboptimal culture conditions are known to increase cellular fragmentation19. Conversely, our data and that of others1,2 supports the notion that the pluripotent inner cell mass in human embryos is more resistant to apoptosis than the trophectoderm, even in karyotypically aberrant cells.

(4) The "count tables generated from the raw data files" could not be found in the source data files.

This slipped to our attention, we have added now the count tables to the source data files. Our apologies.

(5) Citations on aneuploidy literature were not done in a fully scholarly manner. It appears that authors selectively cite previous papers that are in support of their hypothesis but left out those with alternative conclusions.

We apologize if we missed any literature that contradicts our findings, it is not intentional. We would be grateful if the reviewer could provide such references.

In the manuscript we describe the alignment and differences of key findings with several studies (listed below) and the limitations of our study are extensively described in lines 596626.

Our findings align with other work on these aspects:

- RNA-sequencing data2,20–26

- Gene dosage effects drive the transcriptome of the aneuploid human embryo27,28

- Aneuploid cells are cleared by sustained proteotoxic stress followed by p53 activation, autophagy and eventually apoptosis29–37.

- p53 is active in constitutional aneuploid cells38

- The ICM is less sensitive to apoptosis1,2

Our findings differ with other work on these points:

- p53 activation is independent from DNA-damage39

- p53 is active in constitutional aneuploid cells40,41

- Apoptosis is only present in the aneuploid TE of aneuploid cells in the embryo29,30,42

Reviewer #2 (Recommendations For The Authors):

Comments:

(1) The main problem is that there is no substantial novelty. The authors look at previously identified factors affected by chromosome gains and losses, but none of the new one from their analysis. Anything what could be potentially novel is not carefully analyzed (e.g. the difference between reversine-treated and aneuploid samples, or new potential candidates) or explained. This is really a pity.

In the revision, we have further elaborated on the DNA damage aspect by staining for DNA double-stranded breaks and have validated DRAM1 as an activated downstream effector of p53. We have also added the analyses of the gene-expression of the reversine-treated embryos.

(2) Some of the general statements on aneuploidy are confusing and often borderline generalized. E.g. introduction line 106: "If this (proteotoxic stress) remains unresolved by the activation of autophagy..." I am not aware of any publication suggesting that autophagy resolves proteotoxic stress in aneuploid cells. Citations that replication stress causes DNA damage in aneuploid cells are wrong. This link was first shown by Passerini et al. in 2016. etc.

We have clarified these statements in the introduction and added the proposed citations on replication stress that causes DNA damage in aneuploid cells (lines 95-108).

(3) In the figures the authors show a representative image of aneuploid and diploid embryos. Given the aneuploid embryos have widely different karyotypes, it would be important to clarify which of the embryos has been actually shown. Similarly, in the heat maps it is not clear which line is which embryo. This would be very useful.

We added the karyotypes of the aneuploid embryos to the images in figure 3 and 4. Since the heatmaps were removed from the figures we added the karyotypes to the PCAs in all figures.

(4) The authors constantly state that aneuploid embryo accumulate more DNA damage, which is supported by some of their observations, e.g. the DNA damage response is upregulated. It would be great if they would validated this statements with testing some markers for DNA damage.

We agree with the reviewer that this was an important point and addressing it has revealed that our initial assumption was incorrect and has provided new interesting findings. From the revised RNA-seq analysis, we found only one pathway (DNA damage response TP53) to be activated in all aneuploid embryos (Fig.1e). The ATM pathway was also activated specifically in high-dosage embryos. Following this, we set to test if DNA damage was indeed increased in aneuploid embryos by staining for DNA double strand breaks with gH2AX.

First, we investigated the gH2AX expression in 5dpf embryos in which we induced DNAdamage with Bleomycin. We compared 6 untreated versus 6 Bleomycin treated human embryos (Fig. 3m) and found that gH2AX foci were rarely present in the untreated embryos and that all cells of the treated embryos showed a pan-nuclear gH2AX staining.

Second, we compared the presence of gH2AX foci in the TE (NANOG negative cells), ICM (NANOG positive cells) and the whole embryo of 7 euploid versus 11 aneuploid embryos. Interestingly, we found no differences in the number of gH2AX foci or pan-nuclear gH2AX nuclei between euploid and aneuploid embryos (Fig 3o). When dividing our aneuploid embryos into high and low dosage embryos we could also not account for differences. Our data now suggests that complex aneuploid human embryonic cells of meiotic origin do not contain more DNA-double strand breaks, precluding DNA-damage as the source of p53 activation. Last, in our previous experiment we found that phosphorylated S15p53 is increased in aneuploid embryos, supporting an active p53 pathway as suggested by our transcriptomic data. Since we could not find DNA-damage in aneuploid human embryos we speculate that p53 is phosphorylated on Serine15 through metabolic stress as suggested by Jones and colleagues43. We also argue that proteotoxic stress might induce p53 expression as proposed by Singla and colleagues29.

(5) The source of embryos is only partially described in a figure legend. This should be expanded and described in the Materials and Methods section. The embryos are named, but this is nowhere explained. One can only assume that T is for trisomy and M is for monosomy.

We have divided the embryos into different experimental series (Experiment 1-4). This is now described in the Material and methods section (lines 157-175). Also, we have added the experiment number of each embryo to the supplementary tables and to the source data. The abbreviation for T = Trisomy and M= Monosomy was initially introduced in the last paragraph of the figure legend of figure 4. We now added it to every panel.

(6) Recent works from non-embryonic cells suggest that the cellular response to monosomy is different than the response to trisomy. Did the authors try to test this possible difference? For example, one could compare embryos M174/21, M2/19 and M17 with T2/10, T10/22 and T1/15/18/22.

We thank the reviewer for pointing this out. Our RNA-seq. dataset consisted of three embryos that contained trisomies only and four embryos that contained monosomies only. When reanalyzing our data we found different transcriptomic responses between monosomic only and trisomic only cells. Compared to euploid cells, monosomy only cells activate mainly the p53pathway and protein secretion while translation, DNA replication, cell cycle G1/S, DNA synthesis and processing of DNA double strand breaks were inhibited. Trisomy only cells show activated oxidative phosphorylation, ribosome and translation while protein secretion, apoptosis and cell cycle are inhibited. These differences were confirmed by testing transcriptomic differences between trisomic versus monosomic cells. Our results are similar to studies on human embryos20,26 and other monosomic and trisomic cell lines44,45. However, the interpretation of these results is very limited by the small sample size and the comparison of monosomies and trisomies of different chromosomes. Thus, we decided to keep this analysis out of the manuscript.

On the protein level, next to the small sample size, our results were also limited by the fact that not all embryos were stained with the same combinations of antibodies. LC3B was the only protein for which all embryos were immunostained. Thus, other protein data could not be re-analyzed due to even lower sample sizes.

Below we have separated the LC3B puncta per cell counts into euploid, trisomies only, monosomies only and all other aneuploid embryos. We performed a Kruskal Wallis test with multiple comparisons. It is worth noticing that the difference between euploid and monosomies only (and those that contained both) was statistically significant, while the difference between euploid vs trisomies only and trisomies only vs monosomies only was not statistically significant. These differences contradict the studies on monosomic cell lines that found that proteotoxic stress and autophagy are not present and specific to trisomic cell lines. Here we also decided to keep this specific protein expression analysis out of the manuscript due to the above-mentioned limitations.

(7) Line 329: "a trisomy 12 meiotic chromosomal abnormality in one reversine-treated embryo." What does it mean? Why meiotic chromosomal abnormality when the reversine treatment was administered 4 days after fertilization? In the discussion, the authors state "presumed meiotic," but this should be discussed and described more clearly.

Since reversine induces mitotic abnormalities of different types leading to chromosomally mosaic embryos, we could not identify these induced abnormalities using inferCNV on the RNAseq of TE biopsies of said embryos. However, we were not aware of the karyotype of the embryos that were used for these experiments, as they were thawed after they had been cryopreserved at day 3 of development and had not been subjected to genetic testing. This makes it possible that some of those embryos we used for the reversine experiments in fact carried endogenously acquired meiotic and mitotic chromosomal abnormalities. Since we are only able to detect by inferCNV aneuploidies homogeneously present in the majority of the cells of the sequenced biopsy, we only picked up this trisomy 12. It is possible that this was not a meiotic abnormality but a miotic one originating at the first cleavage and present at a high percentage of cells in the blastocyst. At any rate, the exact origin of this aneuploidy has no further implications for the results of the study. We clarified this in the manuscript (lines 310-315).

(8) Line 422: "The gene expression profiles suggest that the accumulation of autophagic proteins in aneuploid embryos is caused by increased autophagic flux due to differential expression of the p53 target gene DNA Damage Regulated Autophagy Modulator-1 (DRAM1), rather than by inhibition of autophagy (Supplementary Table 2)." This is highly speculative, as the authors do not have any evidence to support this statement.

To validate this finding we have now stained 7 euploid and 11 aneuploid embryos with a DRAM1 antibody. We found DRAM1 protein to be significantly enriched in the cytoplasm of TE cells but not in the ICM of aneuploid embryos when comparing with euploid embryos (Fig. 3s,t). This data is consistent with the finding that autophagy is increased in the TE and not the ICM of aneuploid human embryos. (Fig 4l-o). Potential implications of DRAM1 expression have been mentioned in the discussion.

(9) The figure legends are confusing. They are mixed up with the methods and some key information are missing.

We revised all figure legends accordingly and removed the experimental set-up figures from the manuscript to reduce any confusion. The methods section was revised and expanded.

(10) In Figure 1, what is the difference between "activated" and "deregulated"?

Since we analyzed our RNA-seq dataset with the method proposed by reviewer 1 we now generated normalized enrichment scores. The terms activated and deregulated are thus not present anymore.

(11) The p62 images are not really clear. There might be more puncta (not obvious, though), but the staining intensity seems lower in the representative images.

We do not agree with the reviewer that there might be more p62 puncta (purple), however, we agree that it was not clearly visible from the pictures. Below we show an example of the counting mask (in green) of the aneuploid embryo from figure 3i, where one can clearly appreciate that all the puncta are captured by the counting mask. In this case, the software counted 1704 puncta. To further clarify, we now added a zoom of a randomly chose ROI of the p62 staining’s to figure 3i.

(12) The authors claim that there are differences between lineages in response to aneuploidy, such as autophagy not being activated in the OCT4+ lineage, etc. However, the differences are very small and based on a small number of embryos. It is difficult to draw far-reaching conclusions based on a small number of experiments (Fig. 4n-r). The authors also claim in the Abstract that they demonstrated "clear differences with previous findings in the mouse", which are however difficult to identify in the text.

We agree with the reviewer that our conclusions on figures 4l-o were based on a small number of embryos. We have increased as much as possible the sample size. This is challenging due to the constrictions in accessing human embryos, and especially the limited number of embryos with meiotic complex aneuploidy. We have performed immunostainings for LC3B, OCT4 and GATA4 of six additional euploid and four additional aneuploid human embryos. This did not change our overall findings that aneuploid embryos upregulate autophagy in the TE rather than the ICM (Figure 4l-o). After the inclusion of additional embryos, we removed our speculation from the manuscript that autophagy is present in ICM cells of already differentiated cells towards EPI/PrE.

We have rephrased the abstract to state that we highlight a few differences with previous findings in the mouse. Here we focused especially on the different transcriptomic response of reversine treated embryos, that aneuploid mouse embryos do not seem to suffer from lineage segregation errors and that the ICM of aneuploid human embryos lacks apoptosis while aneuploid mouse embryos show elimination from the EPI. Likewise, we highlighted the similar stress responses and that we could give novel insights into p53 mediated autophagy and apoptosis activation through DRAM1 in aneuploid TE cells but not the ICM.

(13) The text needs thorough editing - long sentences, typos, and grammar errors are frequent. Punctuation is largely missing.

We have revised the text.

References

(1) Victor, A. R. et al. One hundred mosaic embryos transferred prospectively in a single clinic: exploring when and why they result in healthy pregnancies. Fertil Steril 111, 280–293 (2019).

(2) Martin, A. et al. Mosaic results after preimplantation genetic testing for aneuploidy may be accompanied by changes in global gene expression. Front Mol Biosci 10, 264 (2023).

(3) Martín, Á. et al. Trophectoderm cells of human mosaic embryos display increased apoptotic levels and impaired differentiation capacity: a molecular clue regarding their reproductive fate? Human Reproduction 39, 709–723 (2024).

(4) Domingo-Muelas, A. et al. Human embryo live imaging reveals nuclear DNA shedding during blastocyst expansion and biopsy. Cell 186, 3166-3181.e18 (2023).

(5) Loewer, A., Karanam, K., Mock, C. & Lahav, G. The p53 response in single cells is linearly correlated to the number of DNA breaks without a distinct threshold. BMC Biol 11, 1–13 (2013).

(6) Kim, H., Watanabe, S., Kitamatsu, M., Watanabe, K. & Ohtsuki, T. Cell cycle dependence of apoptosis photo-triggered using peptide-photosensitizer conjugate. Scientific Reports 2020 10:1 10, 1–8 (2020).

(7) Pollak, N. et al. Cell cycle progression and transmitotic apoptosis resistance promote escape from extrinsic apoptosis. J Cell Sci 134, (2021).

(8) Neufeld, T. P. Autophagy and cell growth--the yin and yang of nutrient responses. J Cell Sci 125, 2359–2368 (2012).

(9) Lanneau, D. et al. Heat shock proteins: essential proteins for apoptosis regulation. J Cell Mol Med 12, 743 (2008).

(10) Gabai, V. L., Mabuchi, K., Mosser, D. D. & Sherman, M. Y. Hsp72 and Stress Kinase cjun N-Terminal Kinase Regulate the Bid-Dependent Pathway in Tumor Necrosis Factor-Induced Apoptosis. Mol Cell Biol 22, 3415 (2002).

(11) Sanchez-Vasquez, E., Bronner, M. E. & Zernicka-Goetz, M. HIF1A contributes to the survival of aneuploid and mosaic pre-implantation embryos. bioRxiv 2023.09.04.556218 (2023) doi:10.1101/2023.09.04.556218.

(12) Chousal, J. N. et al. Molecular profiling of human blastocysts reveals primitive endoderm defects among embryos of decreased implantation potential. Cell Rep 43, (2024).

(13) Corujo-Simon, E., Radley, A. H. & Nichols, J. Evidence implicating sequential commitment of the founder lineages in the human blastocyst by order of hypoblast gene activation. Development (Cambridge) 150, (2023).

(14) Regin, M. et al. Lineage segregation in human pre-implantation embryos is specified by YAP1 and TEAD1. Human Reproduction 38, 1484–1498 (2023).

(15) Saiz, N., Williams, K. M., Seshan, V. E. & Hadjantonakis, A. K. Asynchronous fate decisions by single cells collectively ensure consistent lineage composition in the mouse blastocyst. Nature Communications 2016 7:1 7, 1–14 (2016).

(16) Plusa, B., Piliszek, A., Frankenberg, S., Artus, J. & Hadjantonakis, A. K. Distinct sequential cell behaviours direct primitive endoderm formation in the mouse blastocyst. Development 135, 3081–3091 (2008).

(17) Hashimoto, M. & Sasaki, H. Epiblast Formation by TEAD-YAP-Dependent Expression of Pluripotency Factors and Competitive Elimination of Unspecified Cells. Dev Cell 50, 139-154.e5 (2019).

(18) Hardy, K. Apoptosis in the human embryo. Rev Reprod 4, 125–134 (1999).

(19) Ramos-Ibeas, P. et al. Embryo responses to stress induced by assisted reproductive technologies. Mol Reprod Dev 86, 1292–1306 (2019).

(20) Licciardi, F. et al. Human blastocysts of normal and abnormal karyotypes display distinct transcriptome profiles. Sci Rep 8, 1–9 (2018).

(21) Maxwell, S. M. et al. Investigation of Global Gene Expression of Human Blastocysts Diagnosed as Mosaic using Next-generation Sequencing. Reproductive Sciences 1–11 (2022) doi:10.1007/s43032-022-00899-x.

(22) Groff, A. F. et al. RNA-seq as a tool for evaluating human embryo competence. Genome Res 29, 1705–1718 (2019).

(23) Starostik, M. R., Sosin, O. A. & McCoy, R. C. Single-cell analysis of human embryos reveals diverse patterns of aneuploidy and mosaicism. Genome Res 30, 814–826 (2020).

(24) Vera-Rodriguez, M., Chavez, S. L., Rubio, C., Pera, R. A. R. & Simon, C. Prediction model for aneuploidy in early human embryo development revealed by single-cell analysis. Nat Commun 6, 7601 (2015).

(25) Sanchez-Ribas, I. et al. Transcriptomic behavior of genes associated with chromosome 21 aneuploidies in early embryo development. Fertil Steril 111, 991-1001.e2 (2019).

(26) Fuchs Weizman, N. et al. Towards Improving Embryo Prioritization: Parallel Next Generation Sequencing of DNA and RNA from a Single Trophectoderm Biopsy. Sci Rep 9, 1–11 (2019).

(27) Fernandez Gallardo, E. et al. A multi-omics genome-and-transcriptome single-cell atlas of human preimplantation embryogenesis reveals the cellular and molecular impact of chromosome instability. bioRxiv 2023.03.08.530586 (2023) doi:10.1101/2023.03.08.530586.

(28) Dürrbaum, M. & Storchová, Z. Effects of aneuploidy on gene expression: implications for cancer. FEBS J 283, 791–802 (2016).

(29) Singla, S., Iwamoto-Stohl, L. K., Zhu, M. & Zernicka-Goetz, M. Autophagy-mediated apoptosis eliminates aneuploid cells in a mouse model of chromosome mosaicism. Nat Commun 11, 1–15 (2020).

(30) Bolton, H. et al. Mouse model of chromosome mosaicism reveals lineage-specific depletion of aneuploid cells and normal developmental potential. Nat Commun 7, 1– 12 (2016).

(31) Ohashi, A. et al. Aneuploidy generates proteotoxic stress and DNA damage concurrently with p53-mediated post-mitotic apoptosis in SAC-impaired cells. Nat Commun 6, 1–16 (2015).

(32) Santaguida, S. & Amon, A. Short- and long-term effects of chromosome missegregation and aneuploidy. Nature Reviews Molecular Cell Biology vol. 16 473–485 Preprint at https://doi.org/10.1038/nrm4025 (2015).

(33) Santaguida, S., Vasile, E., White, E. & Amon, A. Aneuploidy-induced cellular stresses limit autophagic degradation. Genes Dev 29, 2010–2021 (2015).

(34) Chunduri, N. K. & Storchová, Z. The diverse consequences of aneuploidy. Nature Cell Biology 2019 21:1 21, 54–62 (2019).

(35) Dürrbaum, M. et al. Unique features of the transcriptional response to model aneuploidy in human cells. BMC Genomics 15, 139 (2014).

(36) Pan, J.-A., Ullman, E., Dou, Z. & Zong, W.-X. Inhibition of protein degradation induces apoptosis through a microtubule-associated protein 1 light chain 3-mediated activation of caspase-8 at intracellular membranes. Mol Cell Biol 31, 3158–70 (2011).

(37) Stingele, S. et al. Global analysis of genome, transcriptome and proteome reveals the response to aneuploidy in human cells. Mol Syst Biol 8, 608 (2012).

(38) Tang, Y.-C., Williams, B. R., Siegel, J. J. & Amon, A. Identification of aneuploidyselective antiproliferation compounds. Cell 144, 499–512 (2011).

(39) Janssen, A., Van Der Burg, M., Szuhai, K., Kops, G. J. P. L. & Medema, R. H. Chromosome segregation errors as a cause of DNA damage and structural chromosome aberrations. Science 333, 1895–1898 (2011).

(40) Li, M. et al. The ATM-p53 pathway suppresses aneuploidy-induced tumorigenesis. Proc Natl Acad Sci U S A 107, 14188–14193 (2010).

(41) Thompson, S. L. & Compton, D. A. Proliferation of aneuploid human cells is limited by a p53-dependent mechanism. J Cell Biol 188, 369–381 (2010).

(42) Yang, M. et al. Depletion of aneuploid cells in human embryos and gastruloids. Nat Cell Biol 23, 314–321 (2021).

(43) Jones, R. G. et al. AMP-activated protein kinase induces a p53-dependent metabolic checkpoint. Mol Cell 18, 283–293 (2005).

(44) Chunduri, N. K., Barthel, K. & Storchova, Z. Consequences of Chromosome Loss: Why Do Cells Need Each Chromosome Twice? Cells 2022, Vol. 11, Page 1530 11, 1530 (2022).

(45) Krivega, M., Stiefel, C. M. & Storchova, Z. Consequences of chromosome gain: A new view on trisomy syndromes. American Journal of Human Genetics vol. 109 2126–2140 Preprint at https://doi.org/10.1016/j.ajhg.2022.10.014 (2022).

Version 2

Abstract

Abstract

About 70% of human cleavage stage embryos show chromosomal mosaicism, falling to 20% in blastocysts. Chromosomally mosaic human blastocysts can implant and lead to healthy new-borns with normal karyotypes. Studies in mouse embryos and human gastruloids have shown that aneuploid cells show proteotoxic stress, autophagy and p53 activation and that they are eliminated from the epiblast by apoptosis while being tolerated in the trophectoderm. These observations suggest a selective loss of aneuploid cells from human embryos, but the underlying mechanisms are not yet fully understood. In this study we investigated the cellular consequences of aneuploidy in a total of 125 human blastocysts. RNA-sequencing of trophectoderm cells showed transcriptional signatures of activated p53 pathway and apoptosis, which was proportionate to the level of chromosomal imbalance. Immunostaining corroborated that aneuploidy triggers proteotoxic stress, autophagy, p53-signalling, and apoptosis independent from DNA damage. Total cell numbers were lower in aneuploid embryos, due to a decline both in trophectoderm and in epiblast/primitive endoderm cell numbers. While lower cell numbers in trophectoderm may be attributed to apoptosis, it appeared that aneuploidy impaired the second lineage segregation, particularly primitive endoderm formation. This might be reinforced by retention of NANOG in aneuploid embryos. Our findings might explain why fully aneuploid embryos fail to further develop and we hypothesize that the same mechanisms lead to removal of aneuploid cells from mosaic embryos. This hypothesis needs further study as we did not analyze chromosomal mosaic embryos. Finally, we demonstrated a few differences with previous findings in the mouse, emphasizing the need for human embryo research to understand the consequences of aneuploidy.

Introduction

Aneuploidy is common in human preimplantation embryos from both natural (Munné et al., 2020a) and medically assisted reproduction cycles (Capalbo et al., 2021; Chavez et al., 2012; Chow et al., 2014; Fragouli et al., 2019; Johnson et al., 2010; Mertzanidou et al., 2013b, 2013a; Popovic et al., 2020; Starostik et al., 2020; Vanneste et al., 2009; Voullaire et al., 2000; Wells and Delhanty, 2000; Yang et al., 2021). Depending on the study, up to 80% of human cleavage stage embryos are found to be aneuploid, of which 70% show chromosomal mosaicism, i.e., the presence of at least two cell lineages with different genomic content (Regin et al., 2022). The reasons behind this high rate of chromosomal abnormalities of mitotic origin is still largely unknown (McCoy, 2017). One explanation is that cleavage-stage embryos have a weak spindle assembly checkpoint (SAC) that is uncoupled of apoptosis until the blastocyst stage (Jacobs et al., 2017; Vázquez-Diez et al., 2019), in combination with an abundancy of transcripts of anti-apoptotic genes (Yan et al., 2013). Only after embryonic genome activation, the rapidly increasing expression of pro-apoptotic genes along with the full functionality of the SAC establishes the control of mitotic errors. Additionally, aneuploidy in human embryos originates from an erroneous first mitotic division through a prolonged prometaphase (Currie et al., 2022).

It is by now well established that chromosomally mosaic embryos can result in healthy new-borns with normal karyotypes, implying a progressive and selective loss of aneuploid cells during development (Capalbo et al., 2021; Fragouli et al., 2019, 2017; Greco et al., 2015; Munné et al., 2020b, 2017; Viotti et al., 2021; Yang et al., 2021). In the preimplantation embryo, several studies have shown that the proportion of aneuploid cells within the embryo decreases as from three days post fertilization (3dpf) (Fragouli et al., 2019; van Echten-Arends et al., 2011; Yang et al., 2021) with no apparent preferential allocation of aneuploid cells to either trophectoderm (TE) or inner cell mass (ICM) (Capalbo et al., 2013; Popovic et al., 2018; Ren et al., 2022; Starostik et al., 2020). Conversely, the TE of mouse embryos treated with the SAC inhibitor reversine show a similar but statistically significant 6% enrichment for aneuploid cells (Bolton et al., 2016). Similarly, other studies on human embryos found that the TE contained a slightly higher number of aneuploid cells than the ICM, but without reaching statistical significance (Griffin et al., 2022; Starostik et al., 2020). Lastly, aneuploid cells have also been found to be excluded as cell debris (Orvieto et al., 2020) or allocated to the blastocoel cavity and to peripheral cells that do not participate in the formation of the embryo (Griffin et al., 2022).

After implantation, aneuploid cells become progressively more frequent in the trophoblast lineage as compared to the epiblast (EPI) and primitive endoderm (PrE) (Starostik et al., 2020). This suggests that the trophoblast cells are more tolerant to aneuploidy, which is in line with the rate of 1-2% of confined placental mosaicism found in the general population (Kalousek and Vekemans, 1996). Large copy number variations appear to be mostly confined to placental lineages (Zamani Esteki et al., 2019) due to developmental bottlenecks occurring during the first lineage segregation which genetically isolate TE from ICM (Coorens et al., 2021). Taken together, there is ample evidence supporting the selective elimination of aneuploid cells from the embryonic lineage during pre- and post-implantation development, while the extraembryonic tissues appear to be more tolerant to genetic imbalances.

From a cellular point of view, aneuploidy has two types of consequences: a generic response and one that depends on the specific aneuploidy. The first type of response is highly conserved and has been found from yeast and plants to mouse and human cells (reviewed in: Santaguida and Amon, 2015; Zhu et al., 2018; Chunduri and Storchová, 2019; Chunduri et al., 2022; Krivega et al., 2022). Aneuploidy results in gene-dosage defects, leading to an unbalanced protein pool that induces the activation of the mechanisms to restore protein homeostasis, such as the chaperone system, the ubiquitin proteasome system and autophagy (Ariyoshi et al., 2016; Dürrbaum et al., 2014; Santaguida et al., 2015; Stingele et al., 2012). If these systems are overwhelmed, the accumulation of misfolded or aggregated proteins induces proteotoxic stress which is detrimental to the cells (Donnelly et al., 2014; Huettel et al., 2008; Ohashi et al., 2015; Oromendia et al., 2012; Stingele et al., 2012; Tang et al., 2011; Torres et al., 2007; Williams et al., 2008). Further, aneuploidy can cause replication stress, which in turn can lead to DNA-damage (Passerini et al., 2016; Santaguida et al., 2017; Sheltzer et al., 2011), p53 activation (Li et al., 2010; Santaguida et al., 2017; Thompson and Compton, 2010) and ultimately to cell cycle arrest and apoptosis (reviewed in: Santaguida and Amon, 2015; Zhu et al., 2018; Regin et al., 2022).

These mechanisms also appear to be active during early embryonic development and to have a lineage specific effect. In mice, clearance of aneuploid cells by apoptosis requires the activation of autophagy and p53 (Singla et al., 2020). Aneuploid cells are only being eliminated by apoptosis in the ICM (Bolton et al., 2016) and at similar rates in the EPI and the PrE (Singla et al., 2020). In a study using human 2D gastruloids with reversine-induced aneuploidy, apoptosis was only induced in differentiating but not in pluripotent cells, although aneuploidy did induce an increase in total nuclear p53 in both cell types. Similarly, as in the mouse, the aneuploid cells were eliminated by apoptosis from the embryonic germ layers but were tolerated in the TE-like cells (Yang et al., 2021).

The type of aneuploidy and the percentage of aneuploid cells within a euploid/aneuploid mosaic embryo determine its developmental capacity. The viability of mouse embryos inversely correlates to the proportion of aneuploid cells, with 50% of euploid cells being sufficient to sustain normal development (Bolton et al., 2016). This same threshold appears to apply to human embryos, where embryos with less than 50% of aneuploid cells have higher developmental potential than those containing more than 50%, albeit both performing worse than fully euploid embryos (Viotti et al., 2021, Capalbo et al., 2021).

Conversely, whole chromosome gains and losses that are present in every cell of the human embryo are usually linked to lethality with the notable exception of trisomies 13 (Patau syndrome), 18 (Edwards syndrome) and 21 (Down syndrome), and sex chromosome abnormalities. The effects of these imbalances can manifest already during early embryo development (Fuchs Weizman et al., 2019; Licciardi et al., 2018; Sanchez-Ribas et al., 2019; Shahbazi et al., 2020). Shahbazi et al. recently described the consequences of carrying specific meiotic aneuploidies on human embryo development using an in vitro implantation model (Shahbazi et al., 2020). They showed that human embryos with trisomy 15, 16 and 21 and monosomy 21 readily all reached the blastocyst stage, and that monosomy 21 embryos arrested after implantation while trisomy 15 and 21 embryos developed further. Interestingly, trisomy 16 embryos showed a hypoproliferative trophoblast due to excessive CDH1 expression, a gene located on chromosome 16.

In this study, we investigated the cellular responses to complex aneuploidy during human preimplantation development, with a focus on stress response and lineage segregation events. We assessed previously unexplored general consequences of naturally acquired aneuploidies in human embryos with focus on impaired lineage segregation events.

Methods

Ethical approval

All experiments have been approved by the local Commission of Medical Ethics of the UZ Brussel (B.U.N. 143201628722) and the Federal Committee for Medical and Scientific Research on Human Embryos in vitro (AdV069 and AdV091). Patients from Brussels IVF (UZ Brussel) donated their embryos for research after written informed consent.

Culture of human pre-implantation embryos

All human embryos, unless stated otherwise, were warmed using the Vitrification Thaw Kit (Vit Kit-Thaw; Irvine Scientific, USA) and cultured in 25 µL droplets of Origio Sequential BlastTM medium (Origio) at 37°C with 5% O2, 6% CO2 and 89% N2. Embryos were graded before vitrification by experienced clinical embryologists according to Gardner and Schoolcraft criteria (Gardner and Schoolcraft, 1999).

Experiment 1

We warmed vitrified human blastocysts (5-6dpf) after PGT (28 euploid and 22 aneuploid) or at 3dpf (n=11). Post PGT-embryos were left to recover for 2h in culture medium, after which they underwent a second TE biopsy. For the reversine experiment, we cultured the 3dpf embryos for 24h in culture medium supplemented with 0.5 µM reversine (Stem Cell Technologies) and after wash-out we let them develop until 5dpf and subsequently took a biopsy. The concentration of 0.5 µM reversine was chosen based on previous experiments in mouse embryos (Bolton et al., 2016; Singla et al., 2020).

Experiment 2

We warmed 12 vitrified and undiagnosed human blastocysts (5-6dpf) of which 6 were cultured in culture medium for 16hrs. The other 6 embryos were cultured in medium supplemented with 0.1mg/mL Bleomycin (Sigma-Aldrich) for 16hrs.

Experiment 3

We warmed vitrified human blastocysts (5-6dpf) after PGT (7 euploid and 11 aneuploid) that were left to recover for 2hrs in culture medium.

Experiment 4

We warmed 5dpf-6dpf PGT embryos (10 euploid and 14 aneuploid) and cultured them for 16h to ensure sufficient time to progress to the second lineage differentiation. Prior to fixation we live-stained all the embryos with either Caspase-3/7 or Caspase-8 (Supplementary Table 9) for 30 min. Additionally, we warmed 5dpf-6dpf PGT embryos (6 euploid and 4 aneuploid) and cultured them for 16h without live-staining.

Biopsy procedure

The zona pellucida was opened (15-25µm) on 4dpf using a Saturn Laser (Cooper Surgical). For PGT cycles, expanded good quality embryos were biopsied on 5dpf; early blastocysts were evaluated again for expansion on 6dpf. Briefly, embryos were transferred to HEPES buffered medium and biopsied on a Nikon inversed microscope equipped with a Saturn Laser. The herniated TE cells were then aspirated and cut using manual laser shoots. In case of sticking cells this was supported by mechanical cutting. After successful tubing, the blastocysts were vitrified using CBS-VIT High-Security straws (CryoBioSystem, L’Aigle, France) with dimethylsulphoxide-ethylene glycol-sucrose as the cryoprotectants (Irvine Scientific Freeze Kit, Newtownmountkennedy, County Wicklow, Ireland) (Van Landuyt et al., 2011). The TE biopsy for RNA-sequencing (Experiment 1) was performed with the same procedure, cells were biopsied from the same opening in the zona pellucida.

PGT

During PGT, our Center of Medical Genetics at the UZ Brussels analyzed a TE sample with multiple cells for each embryo. Next Generation Sequencing: The chromosomal analysis was performed by WGA (Sureplex, Illumina) followed by library preparation (KAPA HyperPlus, Roche), sequencing on a NovaSeq (Illumina) and an in-house developed interpretation pipeline. The analysis has an effective resolution of 5 Mb.

SNP-array

We relied on whole genome amplification (MDA, Repli-G) followed by genome wide SNP array using Karyomapping (Vitrolife) with Bluefuse software (Vitrolife). In addition, SNP array data were analyzed with an in-house developed interpretation pipeline for aneuploidy detection.

RNA-sequencing

Multiple (5-10) TE cells were used as input to generate full length cDNA with the SMART-SeqTM v4 UltraTM Low Input RNA Kit (Clontech Laboratories, Inc.) according to the manufacturer’s instructions. The quality of cDNA was checked using the AATI Fragment Analyzer (Advances Analytical). Library preparation was performed using the KAPA HyperPlus Library Preparation Kit (KAPA Biosystems) according to the manufacturer’s instructions. During cDNA synthesis and library preparation we used 17 and 11 PCR cycles, respectively. Sequencing was performed on a NovaSeq 6000 (Illumina) with 25 million reads per sample.

Bioinformatics analysis

Reads were trimmed using cutadapt version 1.11 to remove the “QuantSEQ FWD” adaptor sequence. We checked the quality of the reads using the FastQC algorithm (Love et al., 2014). Count tables were generated using STAR (Dobin et al., 2013) (version 2.5.3) through mapping against the Genome Reference Consortium Human Build 38 patch release 10 (GRCh38.p10) combined with a general transfer format (GTF) file, both downloaded from the Ensembl database. The RNA-sequencing by Expectation Maximization (RSEM) (Li and Dewey, 2011) software (version 1.3.0) was used to produce the count tables.

Differential gene expression analysis was performed using R-studio (Version 1.1.456) with the edgeR (Robinson et al., 2009) and limma (Ritchie et al., 2015) packages. We included genes with a count per million greater than 1 in at least two samples. The trimmed mean of M values (Robinson and Oshlack, 2010) algorithm was used for normalization. The normalized counts were then transformed in a log2 fold-change (log2FC) table with their associated statistics. In each comparison, genes with a | log2FC | > 1 and an FDR < 0.05 were considered as significantly differentially expressed.

Gene set enrichment analysis (GSEA) was performed using the GSEA software (version 4.3.2) from the Molecular Signature Database (https://www.gsea-msigdb.org/gsea/msigdb/index.jsp). We generated a ranked gene list based on the normalized count matrix of the whole transcriptome that was detected after differential gene expression. The ranked gene list was then subjected to the GSEA function, and we searched the Hallmark and C2 library for significantly enriched pathways. Significance was determined using a cut-off value of 25% FDR. This cut-off is proposed in the user guide of the GSEA (https://www.gsea-msigdb.org/gsea/doc/GSEAUserGuideTEXT.htm) especially for incoherent gene expression datasets, as suggested by our PCAs, which allows for hypothesis driven validation of the dataset.

The copy number variation (CNV) analysis was inferred from our RNA-sequencing data set using inferCNV R package (version 1.7.1) (Inferring copy number alterations from tumor single cell RNA-Seq data (https://github.com/broadinstitute/inferCNV/wiki"). Our count table was used as input and compared to the reference set which in our case was the expression data of the euploid embryos. The analysis followed the standard workflow in inferCNV with the following parameters: “denoise” mode, value of 1 was used for the “cutoff”, prediction of CNV states using default Hidden Markov Models (HMM) method, “cluster_by_groups” was set to TRUE, and other values were set by default.

Gene dosage analysis

We calculated the total number of loci with imbalanced gene expression based on the genetic abnormalities identified during the PGT. For this, we used the number of coding loci per chromosomal region as listed in the ENSMBL database (ensemble.org) and included both gains and losses, either mosaic or homogeneously present, and both full chromosomal aneuploidy as well as segmental abnormalities. The embryos were ranked based on the total number of imbalanced loci and separated into a low- and a high-imbalanced gene dosage group. The low-dosage group contained the embryos with the lowest 50th percentile in number of unbalanced loci, the high-dosage group the embryos with 50th percentile or higher.

Immunocytochemistry

Prior to fixation we removed the zona pellucida using pre-warmed to 37°C Acidic Tyrode’s Solution (Sigma-Aldrich) for 30 sec – 2 min and subsequently washed in PBS/2%BSA. Embryos were then fixed using 4% paraformaldehyde (Sigma-Aldrich) for 10 min, washed three times in PBS/2%BSA for 5 min and then permeabilized using 0.1% Triton X-100 (Sigma-Aldrich) for 20 min followed by another washing cycle. We used PBS/2% BSA supplemented with 10% FBS for 1h to block non-specific protein binding. The embryos were then incubated overnight with primary antibodies (Supplementary Table 9) diluted in blocking buffer at 4°C. The next day, they underwent another washing cycle using PBS/2%BSA and were then incubated with secondary antibodies (1:200, Supplementary Table 9) for 1h at room temperature. After washing, nuclei were stained using Hoechst 33342 (5µg/mL, Life Technologies) diluted in PBS/2%BSA for 15 min. For imaging we mounted the embryos on poly-L-lysine coated glass slides (Sigma-Aldrich) in 3µL of PBS. To avoid flattening of the embryos we used Secure-SealTM Spacers (9mm diameter, 0.12mm deep, Thermofisher) before putting the coverslips in place. For the experiments that required re-staining we imaged the embryos in 3µL PBS/2%BSA in 18 well µ-Slides (Ibidi) and subsequently recuperated them. After recuperation we photobleached the embryos stained for CASP3/7 by using maximum laser power of the 488 nm channel and maximum size of the pinhole for 10 min. The embryos were then re-stained using rat anti-GATA4 primary antibody and the matching Alexa Fluor 488 secondary antibody (Supplementary Table 9) (to replace the CASP3/7 staining).

Imaging and image analysis

Confocal imaging was performed using the LSM800 (Zeiss). Z-stacks were taken with a LD C-Apochromat 40x/1.1 NA water immersion objective in optical sections of 1µm. Nuclei were counted using the Blob finder function of Arivis Vision 4D (3.4.0). All other measurements were performed using the Zen 2 (blue edition) or Image J software based on the optical sections and/or orthogonal projections.

Statistics

The type of statistical test and p-values are mentioned in the figure legends. For most comparisons we used one-way ANOVA or Student’s t-test with or without Welch’s correction. For experiments that contained groups with small sample sizes we used non-parametric tests (Mann-Whitney). Fisher-exact test was used to determine the dependency of the ploidy status on the presence or absence of primitive endoderm. We used GraphPad Prism 9.0.0 for statistical testing. The trend analysis regarding gene dosage effects was performed with SPSS using the Jonckheere-Terpstra test.

Results

Human embryos with complex aneuploidy show gene dosage defects and transcriptomic signatures of p53 activation and apoptosis

We first tested for the presence of a common transcriptomic response to different complex aneuploidies. For this, we studied the transcriptome of a TE biopsy of fifty human blastocysts (5 or 6dpf) previously diagnosed by Preimplantation Genetic Testing (PGT) as either euploid or as containing an aneuploidy of at least two whole chromosomes in all cases except for one embryo (further referred to as aneuploid, detailed karyotypes can be found in the Supplementary Table 1). The aneuploidy was presumed to have occurred during meiosis or during the first cleavage from zygote to the 2-cell stage, resulting in all or most of the cells of the TE biopsy containing the same genetic imbalances. We also included a TE sample of eleven human blastocysts that were treated with 0.5 µM reversine from 3dpf to 4dpf, to induce complex mosaic aneuploidy (referred to as reversine-treated embryos). Of these, 14 euploid, 20 aneuploid and 11 reversine-treated embryos yielded good quality RNA-sequencing results.

Principal component analysis using all expressed genes revealed no clustering of the euploid, aneuploid or reversine-treated embryos into distinct groups (Fig. 1a, Supplementary Figure 1a). We used InferCNV (https://github.com/broadinstitute/inferCNV/wiki) to bioinformatically infer the chromosomal copy numbers based on the higher or lower expression of the genes located on chromosomes with gains or losses (Fig. 1b, Supplementary Fig. 1b,c). We compared the inferred karyotypes to those obtained during PGT and found a match for 45/48 of the full chromosome aneuploidies, while euploidy was correctly predicted in all cases (Supplementary Fig. 1b). These results show that aneuploidy in human blastocysts results in abnormal gene dosage effects. Of note, the karyotypes of the reversine-treated embryos could not be determined since this drug induces different aneuploidies in each cell, leading to mosaic TE biopsies. Also, the karyotypes of the 3dpf embryos were not determined by PGT before the reversine treatment. InferCNV detected only a trisomy 12 in one of the reversine-treated embryos, likely of meiotic or first-cleavage origin (Supplementary Figure 1c).

RNA-sequencing of trophectoderm cells reveals gene dosage defects and transcriptomic signatures of p53 activation and apoptosis

a. Principal component analysis after transcriptome analysis of aneuploid versus euploid embryos. T= Trisomy, M=Monosomy. b. Comparison between the diagnosis of aneuploid embryos obtained after PGT (list on the left) and the results obtained after using InferCNV after RNA-sequencing (plot on the right). Red and blue indicate gain or loss of the chromosome, respectively. c. Volcano plot after differential gene expression analysis with a cutoff value of |log2 fold change| > 1 and -log10(FDR) < 0.05 for aneuploid versus euploid embryos showing 21 upregulated and 2 downregulated genes. d. Barplot of enriched Hallmark pathways after gene set enrichment analysis of aneuploid versus euploid embryos using a cut-off value of 25% FDR. e. Barplot of TOP20 enriched C2 library pathways after gene set enrichment analysis of aneuploid versus euploid embryos using a cut-off value of 25% FDR. f. Enrichment plots of p53-pathway, mitotic spindle, oxidative phosphorylation, MYC targets V1 and ribosome, showing up- or downregulation of genes that are part of the corresponding pathway. Source of all embryos: Experiment 1.

Despite the abnormal gene dosage, but in line with the lack of clear clustering in the PCA, the differential gene expression analysis yielded 21 significantly upregulated and 2 downregulated genes (Fig. 1c, Supplementary Table 2). These findings are similar to those of other studies on the transcriptome of aneuploid human embryonic cells, which also found a few significantly deregulated genes (Fernandez Gallardo et al., 2023; Groff et al., 2019; Martin et al., 2023; Maxwell et al., 2022). We next focused on pathway analysis based on a ranked gene list of the whole transcriptome. We found that cellular stress gene sets were positively enriched in the aneuploid samples, including the p53-pathway and apoptosis (Fig. 1d-f). Gene sets related to metabolism, translation, mitochondrial function and proliferation, such as translation and ribosome, oxidative phosphorylation and MYC targets, were negatively enriched (Fig. 1d-f).

When comparing reversine-treated versus euploid embryos we found 34 significantly upregulated and 22 significantly downregulated genes (Supplementary Fig 1d-f, Supplementary Table 3). The expression profile differed from that identified in the endogenously aneuploid embryos. Here, we found mitotic spindle, WNT-beta catenin signalling, KRAS signalling and lysosome gene sets to be positively enriched while oxidative phosphorylation, MYC targets, translation and ribosome were negatively enriched. We found no evidence for p53 activation or apoptosis. This suggests that naturally acquired aneuploidies of meiotic origin may elicit a different transcriptomic response than reversine-induced mitotic aneuploidies, with a differential induction of the p53 signalling pathway and apoptosis, and inhibition of proliferation.

High gene-dosage imbalances result in high p53 activation and apoptosis while low gene dosage imbalances induce pro-survival pathways

To test whether there is a relationship between the type of cellular stress signatures and the size of the chromosomal imbalances, as previously described in cancer cells (Dürrbaum and Storchová, 2016), we categorized the aneuploid embryos into a low- and a high-imbalanced gene dosage group. This separation was based on the total number of coding loci that were affected by the aneuploidies (Supplementary Table 4). The low-dosage group contained the embryos under the 50th percentile when ranking the embryos from the lowest to the highest number of unbalanced loci. The high-dosage group contained the embryos with 50th percentile or higher. Principal component analysis showed that high-dosage embryos clustered separately from the euploid embryos (Fig. 2a), with 65 significantly upregulated and 153 significantly downregulated genes (Fig. 2b, Supplementary Table 5). High dosage embryos showed a transcriptomic profile consistent with activation of the p53-pathway and apoptosis, while the unfolded protein response, DNA-repair, MYC targets, oxidative phosphorylation, translation and ribosome were inhibited (Fig. 2c-e).

Human embryos with the highest number of genes with abnormal copy number show stronger p53 pathway and apoptosis response.

a. Principal component analysis after transcriptome analysis of high-dosage aneuploid versus euploid embryos. T= Trisomy, M=Monosomy. b. Volcano plot after differential gene expression analysis with a cutoff value of |log2 fold change| > 1 and -log10(FDR) < 0.05 for high-dosage aneuploid versus euploid embryos showing 65 upregulated and 153 downregulated genes. c. Barplot of enriched Hallmark pathways after gene set enrichment analysis of high-dosage aneuploid versus euploid embryos using a cut-off value of 25% FDR. d. Barplot of TOP20 enriched C2 library pathways after gene set enrichment analysis of high-dosage aneuploid versus euploid embryos using a cut-off value of 25% FDR. e. Enrichment plots of p53-pathway, mitotic spindle, unfolded protein response, MYC targets V1 and oxidative phosphorlation, showing up- or downregulation of genes that are part of the corresponding pathway. f. Principal component analysis after transcriptome analysis of low-dosage aneuploid versus euploid embryos. T= Trisomy, M=Monosomy. g. Volcano plot after differential gene expression analysis with a cutoff value of |log2 fold change| > 1 and -log10(FDR) < 0.05 for low-dosage aneuploid versus euploid embryos showing 1 upregulated and 0 downregulated genes. h. Barplot of enriched Hallmark pathways after gene set enrichment analysis of low-dosage aneuploid versus euploid embryos using a cut-off value of 25% FDR. i. Barplot of enriched C2 library pathways after gene set enrichment analysis of low-dosage aneuploid versus euploid embryos using a cut-off value of 25% FDR. j. Enrichment plots of p53-pathway, mitotic spindle, MYC targets V1, unfolded protein response and hypoxia showing upregulation of genes that are part of the corresponding pathway. k. Violin plots with box and whisker plots of the counts per million mapped reads of a supervised set of 10 that are part of apoptosis (BAD, BCL2L1, TNFRSF10B), p53 pathway (DRAM1, MDM2, CDKN1A), and DNA-damage (DDB2). *p = 0.027, **p = 0.010, ***p = 0.004, ****p < 0.001 using the Jonkheere-Terpstra test. Box and whisker plots show median and minimum to maximum values. Source of all embryos: Experiment 1.

In contrast, low-dosage embryos lacked separation from the euploid embryos in the PCA (Fig. 2f) and only had one differentially expressed gene (Fig. 2g). While we still found aneuploidy related stress responses such as p53-pathway and apoptosis to be activated, we found also pro-survival gene sets such as MYC targets (Dang et al., 2006; Wang et al., 2021), MTORC1 signalling (Hung et al., 2012), unfolded protein response (Hetz et al., 2020; Zanetti et al., 2016) and hypoxia (Simões-Sousa et al., 2018) to be enriched (Fig. 2h-j). Low-dosage embryos showed no profiles of ribosome or translation impairment. Last, the plotting of the expression levels of genes in the p53 and apoptosis pathways show their progressive stronger induction or suppression from euploid to low and high dosage embryos (Fig. 2k). This illustrates the association between the severity of the protein pool imbalance and the induction of the cellular stress response.

Aneuploidy triggers proteotoxic stress, DNA-damage independent p53-activation, autophagy and apoptosis

To further investigate the cellular stress responses identified by the transcriptomic analysis, we carried out immunostaining on the same embryos that underwent TE biopsy for transcriptome analysis. The results show that aneuploid and reversine-treated embryos have increased signal intensity of active CASP3/7 and CASP8 (Fig. 3a-d). Both aneuploid and reversine-treated embryos also have lower cell numbers than euploid embryos, likely due to the apoptosis-mediated cell loss (Fig. 3e).

Immunostaining reveals increased proteotoxic stress, DNA-damage independent p53-activation, autophagy and apoptosis

a. Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white) and CASP3/7 (green). T= Trisomy, M=Monosomy. Source: Experiment 1. b. Orthogonal projections after immunostaining of euploid, aneuploid and reversine-treated embryos for DNA (white) and CASP8 (green). T= Trisomy. Source: Experiment 1. c. CASP3/7 mean intensity per cell. Euploid n=13, Aneuploid n=8 embryos. Unpaired t-test, *p = 0.0177. Source: Experiment 1. d. CASP8 mean intensity per cell. Euploid n=6, Aneuploid n=9, Reversine-treated n=8 embryos. One-way ANOVA, *p = 0.04, **p = 0.0098, ns= non-significant. Source: Experiment 1. e. Number of nuclei per embryo. Euploid n=26, Aneuploid n=29, Reversine-treated n=8 embryos. One-way ANOVA, ****p < 0.0001, **p = 0.0033, ns= non-significant. Source: Experiment 1 and 3. f. Orthogonal projections after immunostaining of euploid, aneuploid and reversine-treated embryos for DNA (white), LC3B (turquoise) and HSP70 (magenta). T= Trisomy. Source: Experiment 1. g. LC3B puncta per cell. Euploid n=19, Aneuploid n=22 embryos, Reversine-treated n=8 embryos. One-way ANOVA, **p = 0.0069, ***p = 0.0003, ns= non-significant. Source: Experiment 1. h. HSP70 mean intensity per cell. Euploid n=13, Aneuploid n=8, Reversine-treated n=8 embryos. One-way ANOVA, *p = 0.0121, **p = 0.0096, ns= non-significant. Source: Experiment 1. i. Orthogonal projections after immunostaining of euploid, aneuploid for DNA (white) and p62 (turquoise). White square shows zoom of a representative area. T= Trisomy, M=Monosomy. Source: Experiment 1. j. p62 puncta per cell. Euploid n=6, Aneuploid n=9 embryos. Unpaired t-test, *p = 0.0284. Source: Experiment 1. k. Optical sections of reversine treated embryos for DNA (white), CASP8 (green), HSP70 (red) showing cells with presence (yellow arrow) or absence (yellow arrowhead) of the proteins to investigate co-localization. Source: Experiment 1. l. Percentage of cells of reversine embryos positive for either CASP8/HSP70, CASP8 or HSP70. One-way ANOVA, CASP8/HSP70 vs CASP8***p = 0.0003, CASP8/HSP70 vs HSP70***p = 0.0005, ns= non-significant. Source: Experiment 1. m. Orthogonal projections after immunostaining of euploid, aneuploid for DNA (white) and Serine (S) 15 p53 (turquoise). Source: Experiment 4. n. Percentage of Serine 15 p53 positive cells per embryo. Euploid n = 3, Aneuploid n = 5. Unpaired t-test with Welch’s correction, *p = 0.0356. T= Trisomy. Source: Experiment 4. o. Orthogonal projections of immunostained untreated (n = 6) and Bleomycin treated (n = 6) embryos for DNA (white) and gH2AX (magenta) showing few foci in the untreated and pan-nuclear expression of gH2AX. Source: Experiment: 2. p. Orthogonal projections of immunostained euploid (n = 7) and aneuploid (n = 11) embryos for DNA (white), NANOG (green) and gH2AX (magenta). T= Trisomy. Source: Experiment 3. q. Percentage of cells with gH2AX foci or pan nuclear expression of euploid (n = 7) and aneuploid (n = 11) embryos in NANOG-negative, NANOG positive cells and whole embryos. Source: Experiment 3. r. Percentage of cells with gH2AX foci or pan nuclear expression of euploid (n = 7), low-dosage (n = 5) and high dosage (n = 6) aneuploid embryos in NANOG-negative, NANOG positive cells and whole embryos. Source: Experiment 3. s. Orthogonal projections of immunostained euploid (n = 6) and aneuploid (n = 11) embryos for DNA (white), DRAM1 (turquoise) and NANOG (green). T= Trisomy. Source: Experiment 3. t. DRAM1 mean cytoplasmatic intensity in TE and ICM of euploid and aneuploid cells. Euploid TE (n = 312 cells). Aneuploid TE (n = 434 cells). Euploid ICM (n = 43 cells). Aneuploid ICM (n = 70 cells). Each dot represents the cytoplasm of a single cell. One-way ANOVA, ****p<0.0001, ns = non-significant. Source: Experiment 3.

Embryo sources are indicated in each section. Brightfield pictures were obtained during confocal imaging. All scale bars are 20 µm. Box and whisker plots show median, and whiskers show minimum to maximum values. Bar plots and scatter plots show mean ± s.d.

We next sought to corroborate the transcriptomic signatures of proteotoxic stress and autophagy. The number of puncta per cell of LC3B, and p62, as well as the intensity of HSP70 staining, as markers of autophagy and chronic protein misfolding, were significantly increased in both aneuploid and reversine-treated embryos (Fig. 3f-j). We also tested for the colocalization of CASP8 and HSP70 at the individual cell level. We found that reversine-treated embryonic cells mainly co-express CASP8/HSP70, while the minority is single positive for either CASP8 or HSP70 (Figure 3k,l). These results show that although the transcriptomic profile of naturally occurring aneuploidies and reversine-treated embryos is moderately different (Figure 1c-f, Supplementary Figure 1d-f), the downstream protein response is similar.

One of the key activated pathways identified during the transcriptome analysis was p53-signalling. Thus, we sought to validate p53 activation by staining for the phosphorylation of Serine-15 on p53, which is phosphorylated upon stimuli such as DNA-damage and metabolic stress (Loughery et al., 2014). We found that indeed aneuploid embryos showed a higher percentage of cells with activated p53 (Fig. 3m,n). These differences did not reach significance when considering the ICM/EPI-OCT4-positive cells or TE OCT4-negative cells separately (Supplementary Fig. 2a-c) probably due to insufficient statistical power.

We then stained an additional batch of embryos (Supplementary Table 6) for the double-strand break marker gH2AX, with the aim of determining if the p53 activation was DNA-damage mediated. As a control, we treated human embryos with the DNA-damage inducer Bleomycin and compared them to untreated embryos. All cells of Bleomycin treated embryos showed a pan-nuclear gH2AX pattern, while untreated embryos rarely showed cells containing foci (Fig. 3o). Euploid and aneuploid embryos showed similar fractions of cells with gH2AX foci or pan-nuclear staining (Fig. 3p,q), which did not change when dividing the aneuploid embryos into high and low dosage groups, (Fig. 3r).

Lastly, the transcriptome analysis showed significant upregulation in aneuploid and high-dosage aneuploid embryos of the gene DRAM1 (DNA Damage Regulated Autophagy Modulator-1, Figure 2b). DRAM1 is in the lysosomal membrane, and it increases autophagic flux and apoptosis after stress-induced p53 activation in cancer cells (Crighton et al., 2006; Guan et al., 2015). We stained euploid and aneuploid embryos (Supplementary Table 6) for DRAM1 and NANOG and found that the expression of DRAM1 is increased specifically in the cytoplasm of aneuploid TE cells but not in the ICM (NANOG-positive cells, Figure 3s,t).

Aneuploidy increases apoptosis in the trophectoderm and impairs lineage segregation events

We next investigated whether the stress responses to aneuploidy are cell-type specific, as shown in the mouse (Bolton et al., 2016; Singla et al., 2020) and human 2D gastruloids (Yang et al., 2021). Concretely, we aimed to study the consequences of aneuploidy during the first lineage segregation event to TE and ICM, and the second lineage segregation event to EPI and PrE.

We first immunostained euploid and aneuploid 5-6dpf human embryos for the pluripotency marker NANOG (Supplementary Table 6). NANOG protein was enriched in the nuclei of aneuploid TE and ICM cells (Fig. 4a,b), which could suggest a delay in the exit from pluripotency in both lineages. We then co-stained 6-7dpf euploid and aneuploid blastocysts for OCT4, CASP3/7 and LC3B (Supplementary Table 7). Overall, we found lower cell counts in both the TE (OCT-negative cells) and in the OCT4-positive cells (ICM/EPI cells), Fig. 4c-e) of aneuploid embryos. This resulted in lower total cell numbers in aneuploid embryos (Fig. 4f), in line with our previous observation (Fig. 3e). We found a higher percentage of nuclear CASP3/7 positive cells (Fig. 4g,h) as well as a higher CASP3/7 mean intensity per cell (Fig. 4g,i) specifically in the TE of aneuploid embryos. In contrast, not a single OCT4-positive cell of either the euploid or aneuploid embryos showed nuclear CASP3/7 signal (Fig. 4g). Aneuploid embryos also showed a higher number of CASP3/7-positive micronuclei in the cells (Fig. 4j,k). The lineage-specific response to aneuploidy also extended to presence of autophagy. The staining for LC3B showed that aneuploid embryos displayed overall increased levels of autophagy which was mainly taking place in the TE and not the ICM (Fig. 4l-o).

Aneuploid human embryos show less cells in trophectoderm and OCT4-positive cells and impaired lineage segregation events

a. Orthogonal projections after immunostaining of euploid (n = 7) and aneuploid embryos (n = 10) for DNA (white) and NANOG (green). T= Trisomy. Source: Experiment 3. b. NANOG nuclear enrichment in cells of the TE and ICM of euploid and aneuploid embryos. Each dot represents the nucleus of a single cell. Euploid TE (n = 718 cells). Aneuploid TE (n = 708 cells). Euploid ICM (n = 77 cells). Aneuploid ICM (n = 77 cells). c. Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white) and OCT4 (magenta). The first aneuploid panel (Trisomy 14 and 16) shows a similar number of ICM cells compared to the euploid embryo. The second aneuploid panel (Trisomy 14 and 22) shows an embryo without an ICM. T= Trisomy. Source: Experiment 4. d,e,f. Differences in number of nuclei per embryo between euploid and aneuploid embryos in the (d) OCT4-negative cells (trophectoderm) ***p = 0.0006, (e) OCT4-positive cells (ICM/EPI), **p = 0.0055 and (f) whole embryo ***p = 0.0003. Euploid n = 16, Aneuploid n = 18. Unpaired t-test. Source: Experiment 4. g. Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white), CASP3/7 (green) and OCT4 (magenta). c. T= Trisomy, M= Monosomy. Yellow arrows indicate presence of the signal and yellow arrow heads indicate absence of signal. h. Percentage (%) of CASP3/7 positive (+) cells in the trophectoderm lineage (OCT4-negative cells). Euploid n = 9, Aneuploid n = 14. Unpaired t-test with Welch’s correction, *p = 0.0453. Source: Experiment 4. i. CASP3/7 mean intensity per cell of whole embryos. Euploid n = 9, Aneuploid n = 14. Unpaired t-test with Welch’s correction, *p = 0.0332. Source: Experiment 4. j. Left panel: Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white) and CASP3/7 (green) showing micronuclei. Yellow box indicates zoomed-in region of interest (ROI). Yellow arrows indicate presence of micronuclei and overlap between DNA and CASP3/7. Source: Experiment 4. k, Number of micronuclei per cell in euploid and aneuploid whole embryos. Euploid n = 9, Aneuploid n = 14. Unpaired t-test with Welch’s correction, *p = 0.0115. Source: Experiment 4. l. Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white), LC3B (turquoise) and OCT4 (magenta). Yellow circle indicates the ICM. The first aneuploid panel (Trisomy 1 and 20 and monosomy 18) shows OCT4-positive cells (ICM/EPI) with low levels of autophagy. The second aneuploid panel (Monosomy 14) shows an embryo with high levels of autophagy in the ICM. Source: Experiment 4. m,n,o. Differences in LC3B puncta per cell between euploid (n=10) and aneuploid (n=9) embryos in the (m) whole embryo, *p = 0.0220, (n) OCT4-negative cells (trophectoderm),*p = 0.0172 and (o) OCT4-positive cells (ICM/EPI) ns = non-significant. Mann-Whitney test. Source: Experiment 4. p. Orthogonal projections after immunostaining of euploid and aneuploid embryos for DNA (white), OCT4 (magenta) and GATA4 (green). The first aneuploid panel (trisomy 13 and 15) shows an embryo with presence of GATA4 positive cellc (PrE). The second aneuploid panel (Trisomy 7 and monosomies 5, 12 and 21) shows an embryo that did not contain GATA4 positive cells. Source: Experiment 4. q. Differences in the number of cells per embryo that were OCT4 positive and either GATA4 negative (magenta) or positive (green) or OCT4 negative and GATA4 positive (grey) between euploid (n=15) and aneuploid (n=18) embryos. In case GATA4-positive cells (PrE) were present we considered the GATA4-negative cells to be epiblast (EPI). All euploid embryos contained GATA4-positive cells, 7/18 aneuploid embryos had an ICM completely lacking GATA4-positive cells (Fisher-exact test, p = 0.009). r, Differences in the PrE/EPI ratio between euploid and aneuploid embryos that had an ICM containing GATA4-positive cells. Euploid n = 15, Aneuploid n = 10. Unpaired t-test, **p = 0.0068. Source: Experiment 4. s. Violin plots with box plots of OCT4-positive-cells (ICM/EPI) per embryo (left, ns), OCT4-negative cells (trophectoderm) per embryo (center, ***p < 0.001) and GATA4-positive-cells (PrE) per embryo (right, ***p < 0.001, Jonkheere-Terpstra test) depending on gene dosage. Euploid (light blue), Low dosage (middle blue) and High dosage (darkest blue). Source: Experiment 4.

Brightfield pictures were obtained during confocal imaging. All scale bars are 20 µm. Box and whisker plots show median and minimum to maximum values. Bar plots show mean ± s.d. Besides (b) all dots represent one embryo.

Finally, we observed that the counts of OCT4-positive cells in the aneuploid embryos distributed in two groups: one with very low counts and one with counts in the range of that of euploid embryos (Fig. 4e). This led us to hypothesize that some of the aneuploid embryos had a delayed ICM development, with impairment of the second lineage segregation event. All blastocysts analyzed above were re-stained for GATA4 (PrE marker) and imaged together with OCT4 (ICM/EPI marker). We found that while all euploid embryos contained GATA4-positive cells, 39% of aneuploid embryos had no GATA4-positive cells (Fig. 4p,q). Furthermore, in those aneuploid embryos that contained GATA4-positive PrE cells, there was a significantly lower ratio of GATA4-positive to OCT4-positive/GATA4-negative cells (Fig. 4r). Categorization of the aneuploid embryos into a low- and a high-dosage group (Supplementary Table 8) showed that the higher the number of imbalanced loci, the lower the cell numbers in the TE and PrE but not in the EPI (Fig. 4s).

Taken together, these results suggest that there are important differences in how the embryonic lineages respond to aneuploidy. While the TE is activating autophagy and apoptosis, this is not the case for the ICM/EPI. In turn, while in both lineages, aneuploidy appears to promote a delay in the downregulation of NANOG during differentiation, the ICM of high-dosage aneuploid embryos frequently has poor or no differentiation to PrE.

Discussion

In this work, we studied the consequences of uniform complex aneuploidy in human preimplantation embryos. In our study we find that, while aneuploid embryos had lower cell numbers in both TE and ICM/EPI, the effect of aneuploidy was cell-type dependent. Aneuploid TE presented transcriptomic signatures of p53 signalling and apoptosis, and immunostaining of whole embryos showed that aneuploidy results in increased proteotoxic stress, autophagy, activated p53 independent from DNA-damage and subsequent apoptosis. In contrast, in the ICM, aneuploidy affects PrE formation.

Overall, our gene-expression results are in line with previously published RNA-sequencing studies on aneuploid human embryonic cells that identified transcriptional signatures consistent with an alteration of cell proliferation (Martin et al., 2023; Maxwell et al., 2022; Starostik et al., 2020), the cell cycle (Fuchs Weizman et al., 2019; Licciardi et al., 2018; Martin et al., 2023; Starostik et al., 2020; Vera-Rodriguez et al., 2015), deregulation of autophagy (Licciardi et al., 2018; Sanchez-Ribas et al., 2019), p53 signalling (Licciardi et al., 2018) and apoptosis (Groff et al., 2019; Licciardi et al., 2018; Martin et al., 2023; Maxwell et al., 2022; Sanchez-Ribas et al., 2019).

We found that strength of these transcriptomic alterations is dependent of the number of imbalanced genes. These gene-dosage dependent effects of aneuploidy are akin to those extensively reported in human and mouse aneuploid cancer cell lines (Dürrbaum and Storchová, 2016). We found that higher numbers of aneuploid loci led to stronger induction of p53 target genes and apoptosis responses in the TE cells. Interestingly, high-dosage aneuploid embryos displayed an inhibited unfolded protein response, as also seen in aneuploid embryonic cells by another study (Fernandez Gallardo et al., 2023), while this pathway was activated in low dosage aneuploid embryos. These results suggest that while low-dosage aneuploid cells are trying to restore homeostasis by the unfolded protein response (Hetz et al., 2020; Jäger et al., 2012; Li and Zhu, 2022), high-dosage aneuploid cells are past this point and undergo apoptosis by p53 activation.

In the subsequent experiments, we found that aneuploid embryos showed increased signal of early (CASP8) and late (CASP3/7) caspases, indicating activation of the extrinsic apoptotic pathway (McIlwain et al., 2013). The higher levels of p62 and LC3B we found in these embryos suggests that this activation of apoptosis results at least in part from a sustained interaction between p62, LC3B and CASP8 (Pan et al., 2011). In parallel, aneuploid embryos showed evidence of an activated p53 pathway, not only on a transcriptomic level, but also by presenting elevated levels of Serine 15 phosphorylated p53 (Loughery et al., 2014). This p53 activation has been previously observed on a transcriptomic level in another study on human embryos (Licciardi et al., 2018) and was identified as a key mediator of aneuploidy-dependent apoptosis in the mouse (Singla et al., 2020). While ATM-mediated phosphorylation of p53 on S15 has been reported in cancer cell lines as a consequence of DNA damage during chromosome mis-segregation (Janssen et al., 2011), we found no evidence of increased levels of DNA-damage in our cohort of embryos. It is worth noting that studies on constitutionally aneuploid cell lines have shown contradicting results, with some studies finding p53 activation (Li et al., 2010; Thompson and Compton, 2010) and others not (Tang et al., 2011). An alternative route for p53 activation is through p38, a sensor for cellular stress such as endoplasmic reticulum stress (Mishra and Karande, 2014; Thompson and Compton, 2010). Metabolic alterations might also promote ATM mediated activation of p53 (Li et al., 2010) or phosphorylation on Serine 15 (Jones et al., 2005). Our study also identified DRAM1 as a significantly upregulated gene in aneuploid embryos. DRAM1 is directly regulated by p53 and is not only able to regulate autophagic flux but it also regulates apoptosis, making it likely an important downstream effector of the p53-mediated effects of aneuploidy in human embryos (Crighton et al., 2006; Guan et al., 2015). Lastly, it is worth considering that the complex abnormal human embryonic cells present in our sample might have been subjected to mitotic stress, as suggested by the enrichment of mitotic spindle pathway genes observed in the RNA-seq dataset. Mitotic stress might be induced by sustained proteotoxic stress (Zhu et al., 2018) and can further have contributed to p53 activation. Based on this evidence, we propose that the p53 activation we observed in the embryos with presumed meiotic aneuploidy is akin to that of constitutionally aneuploid cells, and not due to DNA-damage but to cellular stress caused by proteotoxic stress. Taken together, the increased levels of apoptosis we find in aneuploid embryos appear to be mediated at least by two pathways: the sustained p62/LC3B mediated response to unfolded proteins and by p53 activation as a response to cellular stress. An interesting observation from the apoptosis stainings is that aneuploid embryos contained increased numbers of CASP3/7 positive micronuclei. This is similar to findings in a non-human primate model showing that abnormal chromosomes can be encapsulated in micronuclei, which can subsequently be selectively eliminated from the embryo (Daughtry et al., 2019). The presence of DNA in the cytoplasm has also recently been observed during live-imaging of human embryos, which occurs during blastocyst expansion due to nuclear DNA shedding (Domingo-Muelas et al., 2023). In human cleavage stage embryos these micronuclei have been shown to contain genetic material that originates from chromosome breakages due to replication fork stalling and DNA-damage (Palmerola et al., 2022).

In the second part of our work, we focused on studying the lineage-specific response to complex aneuploidy, as the studies on reversine-treated mouse embryos (Bolton et al., 2016; Singla et al., 2020) and human 2D gastruloids (Yang et al., 2021) also identified a cell-type specific effect of aneuploidy. In the mouse, aneuploid cells are eliminated by apoptosis in the EPI, while tolerated in the TE, and in human 2D gastruloids, aneuploid cells are eliminated in the post-gastrulation embryonic germ layers but tolerated in the TE. In the human embryo, the first lineage segregation is established at 5dpf, and the ICM and TE are marked by the differential expression of NANOG and GATA3 (Gerri et al., 2020; Meistermann et al., 2021; Allègre et al., 2022; Regin et al., 2023). We found that both aneuploid ICM and TE cells showed nuclear enrichment of NANOG, which can be suggestive of a delay or impairment of the first lineage segregation event, possibly due to failure or delay in the exit from pluripotency (Zhang et al., 2016).

While we found increased apoptosis in the aneuploid TE, we found that the ICM/EPI had no CASP3/7 positive cells, despite showing lower cell numbers than the euploid ICM/EPI. These results are in line with the findings of Martin and colleagues, where the RNA-sequencing of TE and ICM samples showed disrupted regulation of apoptosis in mosaic aneuploid TE cells while the ICM showed primarily deregulated mitochondrial function (Martin et al., 2023). In contrast, while the study of Victor et al also finds increased apoptosis in the aneuploid TE, some of their embryos did present CASP3/7 positive signals in the ICM. Further research will be needed to understand if the differences with our findings and those of Martin et al are related to technical or to biological factors. A potential explanation for the lack of apoptosis in the human ICM/EPI could be due to an uncoupling between aneuploidy-induced stress signals and apoptosis, as observed in human embryonic stem cells (Mantel et al., 2007). In this context, the lower cell numbers in the aneuploid ICM/EPI could be explained by other known effects of p53 activation next to induction of apoptosis. First, p53 activation can decrease cell proliferation, leading to lower cell numbers. Secondly, p53 activation can downregulate NANOG expression (Abdelalim and Tooyama, 2014; Lin et al., 2005), potentially leading to loss of pluripotency and driving the cells to differentiation and out of the ICM/EPI.

Another observation was that while the aneuploid TE had increased levels of autophagy markers and DRAM1 expression, this did not reach statistical significance in the ICM/EPI. This aligns with work on human embryos showing absence of activation of p53-signalling and autophagy in the transcriptome of the ICM of mosaic human embryos (Martin et al., 2023), but is in contrast with the reversine-treated mouse embryos and human gastruloids, showing increased autophagy and p53 signalling in the ICM lineages (Singla et al., 2020; Yang et al., 2021). Next to a species-specific effect, it is possible that the difference between the mouse, human gastruloid and human studies are because the reversine treatment used had direct effects on p53 (D’Alise et al., 2008; S Santaguida, 2010), leading to the activation of autophagy across the entire cells.

Finally, we found that in those embryos that had already undergone the second lineage segregation event, the aneuploid embryos had a significantly lower ratio of PrE (GATA4-positive) to EPI (OCT4-positive/GATA4-negative) cells than euploid embryos, suggesting that aneuploidy has a negative effect on PrE formation. Interestingly, recent work comparing the gene-expression of euploid and mosaic aneuploid embryos has also found that euploid embryos have a higher expression of the PrE marker SOX17, (Martin et al., 2023).

While in mice, apoptosis is common during PrE formation to eliminate misallocated cells through cell competition (Hashimoto and Sasaki, 2019; Plusa et al., 2008), our data and that of others suggest that the human ICM rarely undergoes apoptosis, even in karyotypically aberrant cells (Martin et al., 2023; Victor et al., 2019; Yang et al., 2021). While a possible explanation suggested by our NANOG stainings is that in aneuploid human embryos the PrE forms later due to retention of NANOG in the ICM and TE, the mechanisms behind this second lineage differentiation defect remain to be elucidated.

When interpreting the results of this study, it is important to bear in mind that we identified the ICM/EPI as staining positive for OCT4 and the PrE as positive for GATA4. We confirmed that the embryos analysed were expanded blastocysts which had undergone the first lineage segregations, as OCT4 was restricted to the inner cells and absent in TE cells (Niakan and Eggan, 2013). The identity of human ICM and EPI cells is currently under vigorous debate (Meistermann et al., 2021; Radley et al., 2022) and exclusive markers of either cell type are untested. Therefore, we cannot assign ICM or EPI identity to OCT4-positive cells with certainty. However, in blastocysts with GATA4-positive cells the second lineage segregation had occurred and PrE was established, and consequently we confidently assigned the EPI identity to OCT4-positive cells in these blastocysts. In GATA4-negative embryos OCT4-positive cells could be either ICM or EPI cells. Interestingly, we only found a single GATA4-positive cell lacking OCT4, supporting the hypothesis that PrE cells differentiate from EPI cells (Meistermann et al., 2021).

Our study has several limitations and there are some considerations to bear in mind. In this work, we did not analyse the transcriptome of the ICM, because the remaining embryos were used for immunostaining after TE biopsy, and we can therefore not draw any conclusions on the ICM transcriptome. We did not study embryos with the same type of aneuploidy and can thus not provide insight on chromosome-specific effects. Although some of our embryos carried genetic abnormalities in mosaic form as part of their complex karyotype (aneuploid/aneuploid mosaics), we did not study diploid/aneuploid mosaic embryos and can therefore not assert that these also activate the same cellular stress pathways in their aneuploid cells. Since these pathways are common to most cell types, our prediction would however be that they are activated in mosaic embryos too. The possibilities for functional studies and lineage tracing experiments in human embryos are very limited, which is why we can only present an observational study. Alternatively, in silico modelling data could be leveraged to address the roles of aneuploidy in blastocyst formation and development (Nissen et al., 2017). The gene-expression analysis revealed that embryos with naturally acquired and induced aneuploidies had few deregulated pathways in common, with mainly the mitotic spindle genes being differentially expressed in both. Conversely, we used reversine to induce aneuploidy in the immunostaining experiments to increase the sample size, and the results showed that the reversine-induced aneuploidies elicited the same stress responses as aneuploidy. This suggests that while reversine-based models are very close, they may not fully mimic endogenous aneuploidy. An alternative is to use the SAC-inhibitor A3146, recently proposed by the Zernicka-Goetz group to induce low-degree aneuploidy in mouse embryos (Sanchez-Vasquez et al., 2023). Finally, our results on the activation of autophagy and proteotoxic stress are also limited by experimental constraints. We cannot rule out indirect activation of apoptosis mediated by autophagy due to the degradation of cell organelles (Gump and Thorburn, 2011). Also, other upstream regulators of autophagy could be involved, such as the cGAS-STING pathway (Krivega et al., 2021). While we and others (Singla et al., 2020) used HSP70 as a marker for chronic protein misfolding, we cannot preclude that is also acts as a (temporary) suppressor of the extrinsic apoptosis pathway in aneuploid cells (Gabai et al., 2002; Lanneau et al., 2008).

Taken together, our results show that while the same pathways appear to be activated in human and mouse embryos with complex aneuploidy, human embryos have a dosage-dependent and differential response when it comes to the downstream effect of these pathways in the different cell lineages. While the mouse readily depletes aneuploid cells by apoptosis specifically from the embryonic lineage before implantation (Bolton et al., 2016; Singla et al., 2020), in the human embryo, apoptosis appears to be more active in the TE and aneuploidy negatively affects PrE formation in the ICM. Our results provide insight into the mechanisms by which human embryos respond to gene-dosage imbalances and may contribute to understanding how aneuploid cells are selectively eliminated in mosaic preimplantation embryos. Further research, especially by using human embryos with endogenous chromosomal mosaicism, will be needed to shed light on the interactions between proteotoxic stress, the p53 pathway and differentiation during early development, for which the most novel models of human embryo implantation (Deglincerti et al., 2016; Kagawa et al., 2021; Shahbazi et al., 2020, 2016) will prove to be of incalculable value.